### 2024-08-06 manuscript_ten-simple-rules-writing-retreat.tex
add ms to repository - still need some edits

% Template for PLoS
% Version 3.6 Aug 2022
%
% % % % % % % % % % % % % % % % % % % % % %
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage[nopatch=eqnum]{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
\usepackage{setspace} 
\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Ten simple rules to organise an effective (student-led) writing retreat}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Nicholas Winterle Daudt\textsuperscript{1,2*},
Claudia Hird\textsuperscript{1,3},
Eleanor Kelly\textsuperscript{1},
Elli Leinikki\textsuperscript{1},
Gretchen McCarthy\textsuperscript{1},
Ian S. Dixon-Anderson\textsuperscript{1},
Jackson Beagley\textsuperscript{1,4},
Jessica Moffitt\textsuperscript{1},
Joseph Curtis\textsuperscript{1},
Lindsay Wickman\textsuperscript{1\textcurrency},
Meghan Duffy\textsuperscript{1,4},
Saskia Foreman\textsuperscript{1},
Leah M. Crowe\textsuperscript{1,2}
\\
\bigskip
\textbf{1} Department of Marine Science, University of Otago, Dunedin, Aotearoa New Zealand
\\
\textbf{2} Department of Mathematics and Statistics, University of Otago, Dunedin, Aotearoa New Zealand
\\
\textbf{3} Department of Zoology, University of Otago, Dunedin, Aotearoa New Zealand
\\
\textbf{4} Department of Geology, University of Otago, Dunedin, Aotearoa New Zealand
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
% \Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
% \ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency Current Address: Marine Mammal Institute, Department of Fisheries, Wildlife and Conservation Sciences, Oregon State University, Newport, OR, USA

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* nicholaswdaudt@gmail.com

\end{flushleft}

% Please keep the abstract below 300 words
%\section*{Abstract}
%Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%\section*{Author summary}
%Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\linenumbers

\section*{Introduction}

Writing is an integral part of academic work. No matter in your career stage, your scholarly output will advance scientific knowledge and count in your career development, getting grants or landing a job. However, in multi-tasking admin, teaching, fieldwork, among other activities, we frequently put aside how important it is to take the proper time to wrap it all up in a written document (e.g. thesis, scientific paper). Scientific writing has its technical tips and tricks, but ultimately, it is a creative act. Yet, the lack of time to think critically, create knowledge and transfer it into words is a significant complaint in academia [~\cite{menzies}]. Therefore, we have to actively disengage from other tasks to enable us to properly engage with the act of writing [~\cite{murray2013}]. In this context, writing retreats have been shown as effective opportunities to unplug from daily work routines, allowing scientists to excel in their passion: producing science.

Writing retreats are structured periods where academics have dedicated time slots for deep-focused writing [~\cite{mcgrail2006, murray2009}]. Initially driven by the motivation to connect with our PhD cohort [~\cite{bernery, omeara}] but also wanting to leverage the moment to have a productive week of work (no pressure, but the clock does not stop ticking!), we set up for a 5-day writing retreat in a remote field station. Once we returned from that week, we realised how much writing we had accomplished and how important it was to share moments as PhDs going through similar experiences in various stages of our degrees. The motivation after that week was such that we organised a second successful retreat edition a year later.

Here, we build on these experiences of student-led writing retreats and lay out Ten Simple Rules for organising an effective retreat. Althuogh the authors are/were all PhD candidates at the time of the retreats, the rules herein should be helpful for research groups and academics more broadly. We cover steps to facilitate the planning/execution of pre- (Rules 1--4), during (Rules 5--9), and pos- (Rule 10) writing retreat actions, but not writing techniques per se. There are many resources on academic writing that we will not delve into but encourage their reading (e.g. [~\cite{turbek2016, hotaling2020, carter2020, cargill2021}]), including articles in this series [~\cite{weinberger2015, mensh2017, peterson2018}]. Otherwise, you may consider writing a boring contribution instead [~\cite{sand2007}].

\section*{Rule 1: Leverage university facilities}

A change of scenery can help inspire productivity whether the writing retreat is structured as a locally organized event or as multi-day trip. Local facilities are often more straightforward as this can include meeting rooms and other bookable campus spaces. For longer retreats, students, regardless of their departmental affiliation, should inquire about their university’s field stations. Many biology and ecology programmes, for example, have field stations (for a global list, see [~\cite{tydecks}]), and by leveraging university owned facilities, retreat costs can be kept within a manageable budget. There are often times of the year when field stations are underutilized, and having another reason for people to be in a field station during these periods can help with ongoing upkeep. In addition, field stations are often in remote, beautiful settings amongst nature with limited distractions providing an environment conducive to focusing and connecting with peers. If your university doesn’t have a field station, or one that would properly support a writing retreat, other universities are often open to shared use if availability allows. We encourage students in this situation to inquire with station managers to build those relationships.

When determining if a space is suitable for a writing retreat, it is important to take an inventory of available amenities both within the facility and in the area. We made use of our department’s marine field station in the town of Oban on Rakiura/Stewart Island, a small island off the southern coast of the South Island of Aotearoa New Zealand. This location suited us well as everything we needed was within walking distance. Because this is a small community (around 400 year-round residents) abutting a national forest, we had limited distractions, straightforward options for restaurants and groceries, and we had ample options for walking tracks and beach strolls. This field station is a three-bedroom home with twelve bunks, three bathrooms, a kitchen, and an open space with a large table, two desks, and two couches. Ten people attended in each of the two years which was a comfortable number for the space. When determining if a space will work for a retreat, it is also important to understand the internet availability, the heating/cooling systems, sleeping spaces, and work spaces (including available power points).

\section*{Rule 2: Prepare a proposal}

Before you reach out for support, it is important to be prepared and put together a proposal. Your proposal should include possible locations, accommodation, transport, dates and potential participants. Deciding how many people and specifically who the retreat is for is crucial as it is required for other planning details. Numbers may be decided for you based on the accommodation available (see Rule 1) or it could be the other way around. Once you have outlined possible locations and numbers, it is important to work out associated costs. Additional considerations may include transportation to, from and around the writing retreat location, food requirements, and dates/times.

Once you have written your proposal, you can reach out for support. The first port of call should be the departments of the students attending the retreat. Other bodies within the university may be able to fund and support your writing retreat such as higher divisions and postgraduate support offices. Industry, community organisations (e.g. Lions and Rotary Clubs) and fundraising may be able to provide supplemental support, if needed. A well-considered proposal will provide a good foundation for your retreat and set you up for success.

\section*{Rule 3: Structure your retreat}

We were fortunate that our department was very supportive of the writing retreat proposal, and when planning the first trip, encouraged us to use experts on staff at the university. At the University of Otago, we worked with the Higher Education Development Centre which provided resources and guidance on how to structure both productive days and the overall week. There are many resources available to guide writing retreats for PhD students, including this one, but we encourage student to take advantage of these kinds of services at their universities if they are available.

Our week was organized so that we traveled to and from the field station on Sunday and Saturday to allow for Monday through Friday to be fully structured for the retreat. We left extra time on travel days to settle in/pack up, grocery shop, and explore the area. Upon arrival, it was important to have a short welcome discussion to establish ground rules and make sure everyone was comfortable in the space (see Rule 5). Each work day was themed to give some guidance on what to try and tackle at that point in the week (Table~\ref{table1}). We also made sure to intersperse optional fun activities throughout the week which included a pub quiz (i.e. bar trivia), wildlife tours, game night, and other local events.

% Place tables after the first paragraph in which they are cited.
\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Daily themes over a 5-day writing retreat.}}
\begin{tabular}{p{1in}p{2in}p{2in}}
\hline
{\bf Weekday} & {\bf Theme} & {\bf Description}\\ \thickhline
Monday & Getting started & Commit to the goal for the week (see Rule 4), organize a strategy, and establish accountability partners.\\ \hline
Tuesday & Words on the page & Just start writing something! (Rule 6)\\ \hline
Wednesday & Keep going! & Continue writing and refine. (Rule 6)\\ \hline
Thursday & Good enough & Exchange work within accountability partners to practice giving and receiving comments. Use this opportunity to reflect on writing structure and style from a reader’s perspective. (Rule 8)\\ \hline
Friday & Wrap it up & End the retreat by reflecting on what was accomplished this week (Rule 10) and develop a plan on how to proceed.\\ \hline
\end{tabular}
\label{table1}
%\end{adjustwidth}
\end{table}

The schedule for each day was roughly between 9 am and 6 pm and was provided in advance of the retreat so participants knew what to expect for dedicated work time (see \nameref{S1_Appendix}). A different leader was designated for each day to lead discussions and keep the time for writing blocks (between 60 and 120 min) and breaks (between 30 and 60 min). We started each morning together by writing for 15 minutes using prompts that varied from reflections on the different ways we write in our lives to the most ridiculous title we could conjure for our research. This was a productive way to get our writing engine revving and a fun opportunity to share writing with each other. Each evening we ended the day by a reflective discussion on how the day went (with an 'accountability buddy' [e.g. 'Rule 5' in ~\cite{peterson2018}]), a shared dinner, and presentations (see Rule 7).

\section*{Rule 4: Have a pre-retreat meeting}

Hosting a pre-retreat meeting with attendees before departure facilitates a smoother and more productive retreat by addressing the primary objectives of introductions, finalizing logistics, and goal-setting. During the event, participants will get a chance to meet (if they haven’t already) and establish or reinforce a familiar group dynamic, an essential prerequisite for a retreat atmosphere of respect and accountability (see Rules 5 and 10). Fine-tuning the plan for retreat logistics and scheduling through group conversation can identify opportunities for clarification and improvement, and create a sense of shared ownership over day-to-day programming. Most importantly, a pre-retreat meeting provides a chance for participants to identify or workshop specific writing goals with time to prepare relevant resources (literature, data analysis, input of academic supervisors) so that these non-writing activities don't become distractions during the initial days of the retreat.

We suggest participants should arrive at the pre-retreat meeting with a preliminary goal to share with the group. Goals can be quantitative (word count, page number) or qualitative (paper sections, revision of existing work). For reference, the average output during our 5-day retreats was 1.2 pages or 550 words per day, and ranged from refined work on a single chapter/manuscript section to drafts of entire manuscripts. Critically, the specific thesis component (or other written work) should be identified at the meeting to remove the friction of choosing a piece of writing at the outset of the retreat and maximizing preparation time. Sharing specific goals has the additional benefits of introducing the group to everybody's topic of research, as well as identifying whether a participant's preliminary goal is appropriately ambitious for an extended period of dedicated writing.

We recommend that a pre-retreat meeting be held no more than two weeks before departure, allowing time for goal-oriented preparation while limiting stagnation of attention and momentum heading into the retreat. If possible, hosting the event in person (ideally off-campus) is preferrable, although most of the aims of the meeting could be achieved via video conference if necessary (for instance, if you have people from other campi attending the retreat). Secondary objectives for a pre-retreat meeting could include selecting creative writing prompts, topics for presentations (if these are included in your retreat), "get to know you" activities, signing up for roles (meal spots, daily facilitator), or crowd-sourcing ideas for social aspects such as field trips or evening activities. If a planning event cannot be arranged before the retreat, objectives of the meeting (introductions, goal-setting, flexible logistic planning) should still be performed during the first day of the retreat (likely just after discussing the ground rules; see Rule 5), though this will come at the expense of writing time, momentum, and opportunities for preparation.

\section*{Rule 5: Establish ground rules}

At the start of your retreat, it is important to establish clear ground rules and expectations. These can include shared expectations for the group, such as setting quiet hours, assigning household responsibilities (i.e. cooking and cleaning) and setting guidelines for breaks. The document used as the basis for our ground rules can be found in \nameref{S2_Appendix}. Ground rules should also set the tone of the retreat, creating a balanced foundation for respect, productivity and enthusiasm.

At our retreat, the primary guideline we established was a focus on respectful and non-judgemental interactions between all participants. This attitude allowed flexibility in goal setting and accomplishments depending on each person's needs and stage of their PhD. The baseline expectation of respect also led to creating clear expectations for working versus quiet hours and maintaining a productive working environment. During our ground rule meeting, we also discussed the need for flexibility to ensure our retreat catered to the needs of all the participants. For example, not all participants are native English speakers and may find full days of writing and discussion in English more tiring. As a result, we designated certain writing blocks as optional---if participants needed a longer break or felt their time would be better spent refreshing their mind with a forest walk or gym session, there was no judgement. Continuing with the guideline of respect, our opening meeting also made it clear that while group discussions and feedback are useful, only constructive criticism would be tolerated. Together, clear expectations, a set schedule and overall attitude of respect set the tone for a successful week of writing.

\section*{Rule 6: Write}

This is why you wanted to organise a retreat in the first place---to make progress on your academic writing. You have already created a schedule and a daily theme for yourself (see Rule 3) so it’s time to follow it. This is where the goals that you have set for yourself (see Rule 4) also come into play. Knowing what you are working towards, and having a plan of how you intend to get there, should provide you with the space needed to put words on the page and have productive writing blocks. However, writing is not easy and writers face many hurdles[~\cite{grogan2021}], such as a lack of motivation, uncertainty about what to do next, writer’s block, imposter syndrome, and more. These obstacles can easily disrupt even the best-laid plans. It is therefore important to have a backup plan.

If you are struggling to get any writing done it may be a good time to take a break or switch tasks. It is proven that even short breaks can increase productivity [~\cite{lyubykh, carter2020}]. Alternatively, this could be a good time for you to do some research into writing development. If you are not being productive in writing, you may as well be productive in learning how to write. Everybody has different strategies that work for them. For example, some people prefer to cite as they write, whereas others prefer to write everything out quickly and then add citations as they edit. Some authors, such as Ernest Hemingway [~\cite{hemingway}], have even claimed that they purposely leave a sentence, paragraph, or idea unfinished at the end of their writing day specifically so they have something easy to start with for their next writing block. It is important to understand what works for you in terms of a writing practice and to reevaluate your process often to ensure it is still working [~\cite{peterson2018, grogan2021}].

\section*{Rule 7: Lead an academic discussion}

While the primary objective of a writing retreat is to write, there are also opportunities for other forms of professional development. Undertaking the PhD journey is daunting and different for every student, and it is easy to become so engrossed in your own work, that you forgot you are surrounded by people in the same situation as you. A writing retreat provides the opportunity to share your own research or discuss scientific/academic topics amongst your peers. While we found this occurs naturally throughout the retreat, especially within groups of similarly focused researchers, it was decided that a small block of time each day (see Rule 3) would be specifically allocated for students to share a talk or lead a discussion of their own with the other attendees. The casual, non-judgemental environment gave presenting students a chance to practice a seminar talk, get insight on methodology, analysis and results, or simply share "tips and tools of the trade". For many students, this was an opportunity to learn about the research their peers conduct; in a field as multi-faceted as marine science, it is common that we are unaware of the scope or type of research our peers are involved in.

We suggest scheduling the discussions after writing blocks are complete for the day and spread them out over the retreat. Access to a projector and screen in our second retreat was beneficial. However, previously, we have simply shared from our laptop screens or consulted notes and found it just as engaging due to the small group attending the retreat. Talks were not strictly timed as it allowed for conversation throughout, but we aimed not to exceed 20 min per discussion. It is important to emphasise the casualness of such an exercise---students should not be using their time for writing anxiously preparing a conference-level presentation. It is simply an opportunity to share, learn, provide and gain insight. Importantly, these sections give a chance get to know one another and bound as fellow PhD students [~\cite{bernery, omeara}].

\section*{Rule 8: Review and be reviewed}

An important aspect of writing is the practice of giving and receiving revisions with your peers. It is important to remember that the feedback exchanged between peers should be constructive and can offer insight on gaps in clarity, logic, or structure. The use of neutral language should be encouraged [~\cite{parsons2021}]. There is the possibility you may not understand content your peer is working on, however, focusing on aspects of structure, prose, and overall writing flow is a good place to start with peer review and revisions. Receiving feedback from a peer outside of their field may provide different perspectives the author had not previously considered and can lead to more balanced and comprehensive writing. It is also important to receive feedback as much as you can provide it as this practice will encourage a collaborative approach to writing which can build confidence and resilience in both the reviewed and the reviewer.

\section*{Rule 9: Have fun}

A key aspect of a writing retreat is to have fun and enjoy the whole experience---from writing to exploring the local area. While on a writing retreat, it is important to take opportunities to experience new activities you may not usually get to. A retreat also presents a unique chance to bond with your peers while balancing downtime activities and writing time for optimal productivity.

Before you embark on your writing retreat, we recommend that you do some research before arriving to see what is unique about the area to take advantage of what may be on offer. While based at our university’s field station, we had opportunities to aurora hunt and kiwi spot, explore the nearby Ulva island (a protected area), participate in bar trivia, and develop relationships with some of the locals. We also organised creative morning writing prompts (see Rule 3) and gave casual presentations (Rule 7) about various topics which were fun and engaging bonding activities.

More rural places often have scenic landscapes, hospitable communities and limited distractions. This could restrict activity choices but may produce more quality experiences where you can get to know the local businesses and people on a more personal level. A retreat is not only about writing but an opportunity to engage with peers and have fun.

\section*{Rule 10: Posterity -- Be accountable and acknowledge}

Gathering feedback from participants of the retreat and drafting a report are important actions following the writing retreat. Allowing time for the participants to reflect on and discuss to their peers the pitfalls and peaks of the retreat on the last day or two of the retreat and then having them fill out a anonymous survey immediately after returning is very beneficial. By summarizing these points into a concise report soon after attending the retreat, points are clear in the minds of the participants and key issues or takeaways are not forgotten. Comments of the schedule, notes on things that went well, and notes of things that might need improvements are valuable to cement ideas and details that will fade in time before another retreat can be planned. This will reinforce great points from the retreat but more importantly creates some accountability for posterity. Knowledge transfer from one retreat to the next is difficult in an ever-changing student body, but creating a paper trail of advice and resources within your own department provides blueprints and templates for new volunteers to organize their own retreats. 
%The post-retreat report can act as a manual for future students and helps to lower the barrier to organizing a productive event by being a reference to future leaders with answers to FAQs and examples of how to organize administrative tasks.

In addition, give acknowledgement to the writing retreat in your work and publications and circulate this knowledge to your peers, your department, and your colleagues. Having acknowledgements of the retreat (potentially with a uniform phrase) can even provide a searchable tool in online databases. Creating a repository where work that has benefited from the writing retreat can be achieved by filtering these acknowledgements and compiling publications together in the same place. Highlighting to your department the quantity and quality of work that results from a writing retreat is key to securing the longevity of writing retreats. A list which updates into the future as new work is published that stemmed from the writing retreat means that administrative personnel can become aware of the significant impact this opportunity has provided student researchers. Both the scientific community and the university administration often use publication numbers as their tangible metric to interpret how their financial decisions have made an impact. Evidence of the how many papers may come out of a writing retreat by acknowledging the usefulness of it will help to capitalizing on this trend. A scientifically-productive retreat may ease the justification process of future student organizers, or even incentivise greater financial support from administrative groups if the expenditure of writing retreats are seen as a net-positive for publications.

\section*{Conclusion}

Following these set of rules, our writing retreats were sucessful. Colectivelly, we indeed experienced benefits reported in the literature, such as feeling empowered writers through supporting each other in a respectful and non-judgemental environment [~\cite{papen2018}], increasing our sense of belonging as PhD students [~\cite{omeara}] and decreasing our sense of isolation [~\cite{eardley2021}]. However, most importantly, we got our writing reatreat goals achieved. Some of the references herein can help with the theoretical side of writing, but ultimatelly writing is practice [~\cite{peterson2018, grogan2021}]. By facilitating the planning and execution of writing retreats, we hope this piece can foster scholarly productivity. Happy writing!

\section*{Supporting information}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Appendix.}
\label{S1_Appendix}
{\bf Schedule Example.} Brief description...

\paragraph*{S2 Appendix.}
\label{S2_Appendix}
{\bf Ground Rules.} Brief description...

\section*{Acknowledgments}
This manuscript is a contribution conceptualized, led, and written by PhD students, who attended writing retreats supported by the Department of Marine Science of the University of Otago, Aotearoa New Zealand. We hope writing retreats stay as a trend, fostering collaboration and friendship among students and promoting quality science. Thanks also to Tracy Rogers and Claire Hudson from the Higher Education Development Centre of the University of Otago for the support while organising both writing retreats, and the Southern Sea Hotel and the Rakiura/Stewart Island community for welcoming us.

\section*{Author Contributions}

\paragraph*{Conceptualization:} Nicholas W. Daudt (lead); Claudia Hird; Eleanor Kelly; Elli Leinikki; Gretchen McCarthy; Ian S. Dixon-Anderson; Jackson Beagley; Jessica Moffitt; Joseph Curtis; Lindsay Wickman; Meghan Duffy; Saskia Foreman; Leah M. Crowe.

\paragraph*{Project administration:} Nicholas W. Daudt.

\paragraph*{Writing – original draft:}  Nicholas W. Daudt (lead); Claudia Hird; Eleanor Kelly; Gretchen McCarthy; Ian S. Dixon-Anderson; Jackson Beagley; Joseph Curtis; Meghan Duffy; Saskia Foreman; Leah M. Crowe.

\paragraph*{Writing – review & editing:}  Nicholas W. Daudt (lead); Claudia Hird; Eleanor Kelly; Elli Leinikki; Gretchen McCarthy; Ian S. Dixon-Anderson; Jackson Beagley; Jessica Moffitt; Joseph Curtis; Lindsay Wickman; Meghan Duffy; Saskia Foreman; Leah M. Crowe.

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
% 
%\begin{thebibliography}{10}

%\bibitem{tydecks}
%Tydecks L, Bremerich V, Jentschke I, Likens GE, Tockner K.
%\newblock {{B}iological field stations: {A} global infrastructure for research, education, and public engagement}.
%\newblock BioScience 2016;66(2):164--171.

%\bibitem{hemingway}
%Hemingway E.
%\newblock {{M}onologue to the {M}aestro: {A} high seas letter}.
%\newblock Esquire, Oct; 1935.

%\bibitem{lyubykh}
%Lyubykh Z, Gulseren D, Premji Z, Wingate TG, Deng C, Bélanger LJ, Turner N.
%\newblock {{R}ole of work breaks in well-being and performance: {A} systematic review and future research agenda}.
%\newblock J. Occup. Health Psychol. 2022;27(5):470--487.

%\end{thebibliography}

\bibliography{references}

\end{document}

### 2024-08-07 manuscript_ten-simple-rules-writing-retreat.tex
finalised the first draft

@@
 Writing is an integral part of academic work.
-No matter in your career stage, your scholarly output will advance scientific knowledge and count in your career development, getting grants or landing a job.
+No matter in what career stage you are at, your scholarly output will advance scientific knowledge and count in your career development, getting grants or landing a job.
 However, in multi-tasking admin, teaching, fieldwork, among other activities, we frequently put aside how important it is to take the proper time to wrap it all up in a written document (e.g.
@@
 Althuogh the authors are/were all PhD candidates at the time of the retreats, the rules herein should be helpful for research groups and academics more broadly.
-We cover steps to facilitate the planning/execution of pre- (Rules 1--4), during (Rules 5--9), and pos- (Rule 10) writing retreat actions, but not writing techniques per se.
+We cover steps to facilitate the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and pos- (Rule 10) writing retreat actions, but not writing techniques per se.
 There are many resources on academic writing that we will not delve into but encourage their reading (e.g.
@@
 The schedule for each day was roughly between 9 am and 6 pm and was provided in advance of the retreat so participants knew what to expect for dedicated work time (see \nameref{S1_Appendix}).
-A different leader was designated for each day to lead discussions and keep the time for writing blocks (between 60 and 120 min) and breaks (between 30 and 60 min).
+A different facilitator was designated for each day to lead discussions and keep the time for writing blocks (between 60 and 120 min) and breaks (between 30 and 60 min).
 We started each morning together by writing for 15 minutes using prompts that varied from reflections on the different ways we write in our lives to the most ridiculous title we could conjure for our research.
 This was a productive way to get our writing engine revving and a fun opportunity to share writing with each other.
-Each evening we ended the day by a reflective discussion on how the day went (with an 'accountability buddy' [e.g.
-'Rule 5' in ~\cite{peterson2018}]), a shared dinner, and presentations (see Rule 7).
+Each evening we ended the day by a reflective discussion on how the day went (with an 'accountability buddy' [see Rule 5 in ~\cite{peterson2018}]), a shared dinner, and presentations (see Rule 7).
 
@@
 Hosting a pre-retreat meeting with attendees before departure facilitates a smoother and more productive retreat by addressing the primary objectives of introductions, finalizing logistics, and goal-setting.
-During the event, participants will get a chance to meet (if they haven’t already) and establish or reinforce a familiar group dynamic, an essential prerequisite for a retreat atmosphere of respect and accountability (see Rules 5 and 10).
+During the event, participants will get a chance to meet (if they haven't already) and establish or reinforce a familiar group dynamic, an essential prerequisite for a retreat atmosphere of respect and accountability (see Rules 5 and 10).
 Fine-tuning the plan for retreat logistics and scheduling through group conversation can identify opportunities for clarification and improvement, and create a sense of shared ownership over day-to-day programming.
-Most importantly, a pre-retreat meeting provides a chance for participants to identify or workshop specific writing goals with time to prepare relevant resources (literature, data analysis, input of academic supervisors) so that these non-writing activities don't become distractions during the initial days of the retreat.
+Most importantly, a pre-retreat meeting provides a chance for participants to identify or workshop specific writing goals with time to prepare relevant resources (literature, data analysis, input of academic supervisors) so that these non-writing activities do not become distractions during the initial days of the retreat.
 
@@
 
-A key aspect of a writing retreat is to have fun and enjoy the whole experience---from writing to exploring the local area.
-While on a writing retreat, it is important to take opportunities to experience new activities you may not usually get to.
-A retreat also presents a unique chance to bond with your peers while balancing downtime activities and writing time for optimal productivity.
-
-Before you embark on your writing retreat, we recommend that you do some research before arriving to see what is unique about the area to take advantage of what may be on offer.
-While based at our university’s field station, we had opportunities to aurora hunt and kiwi spot, explore the nearby Ulva island (a protected area), participate in bar trivia, and develop relationships with some of the locals.
-We also organised creative morning writing prompts (see Rule 3) and gave casual presentations (Rule 7) about various topics which were fun and engaging bonding activities.
-
+A key aspect of a writing retreat is to have fun and enjoy the whole experience.
+While on a writing retreat, you might as well experience new activities you may not usually get to.
+A retreat also presents a unique chance to bond with your peers while balancing downtime activities and writing time for optimal productivity [~\cite{eardley2021, lyubykh}].
 More rural places often have scenic landscapes, hospitable communities and limited distractions.
@@
 
-\section*{Rule 10: Posterity -- Be accountable and acknowledge}
+Before you embark on your writing retreat, we recommend that you do some research before arriving to see what is unique about the area to take advantage of what may be on offer.
+While based at our university's field station, we had opportunities to aurora hunt and kiwi spot, explore the nearby Ulva Island (a protected area), participate in bar trivia, and develop relationships with some of the locals.
+We also organised creative morning writing prompts (see Rule 3) and gave casual presentations (Rule 7) about various topics which were fun and engaging bonding activities.
+
+\section*{Rule 10: Posterity---be accountable and acknowledge}
 
@@
 This will reinforce great points from the retreat but more importantly creates some accountability for posterity.
-Knowledge transfer from one retreat to the next is difficult in an ever-changing student body, but creating a paper trail of advice and resources within your own department provides blueprints and templates for new volunteers to organize their own retreats.
+Knowledge transfer from one retreat to the next is difficult in an ever-changing student body, but creating a paper trail of advice and resources within your own department provides blueprints and templates for new volunteers to organise their own retreats.
+
 %The post-retreat report can act as a manual for future students and helps to lower the barrier to organizing a productive event by being a reference to future leaders with answers to FAQs and examples of how to organize administrative tasks.
+-- not sure if we did this, so maybe better to leave it out?
 
 In addition, give acknowledgement to the writing retreat in your work and publications and circulate this knowledge to your peers, your department, and your colleagues.
-Having acknowledgements of the retreat (potentially with a uniform phrase) can even provide a searchable tool in online databases.
-Creating a repository where work that has benefited from the writing retreat can be achieved by filtering these acknowledgements and compiling publications together in the same place.
+\textcolor{red}{Having acknowledgements of the retreat (potentially with a uniform phrase) can even provide a searchable tool in online databases.} Creating a repository where work that has benefited from the writing retreat can be achieved by filtering these acknowledgements and compiling publications together in the same place.
 Highlighting to your department the quantity and quality of work that results from a writing retreat is key to securing the longevity of writing retreats.
 A list which updates into the future as new work is published that stemmed from the writing retreat means that administrative personnel can become aware of the significant impact this opportunity has provided student researchers.
-Both the scientific community and the university administration often use publication numbers as their tangible metric to interpret how their financial decisions have made an impact.
+Both the scientific community and the university administration often use publication numbers as their metric to interpret how their financial decisions have made an impact.
 Evidence of the how many papers may come out of a writing retreat by acknowledging the usefulness of it will help to capitalizing on this trend.
-A scientifically-productive retreat may ease the justification process of future student organizers, or even incentivise greater financial support from administrative groups if the expenditure of writing retreats are seen as a net-positive for publications.
+A scientifically-productive retreat may ease the justification process of future student organisers, or even incentivise greater financial support from administrative groups if the expenditure of writing retreats are seen as a net-positive for publications.
 
@@
 
-Following these set of rules, our writing retreats were sucessful.
-Colectivelly, we indeed experienced benefits reported in the literature, such as feeling empowered writers through supporting each other in a respectful and non-judgemental environment [~\cite{papen2018}], increasing our sense of belonging as PhD students [~\cite{omeara}] and decreasing our sense of isolation [~\cite{eardley2021}].
-However, most importantly, we got our writing reatreat goals achieved.
+Following these set of rules, our  were sucessful.
+
+Colectivelly, we experienced the benefits of writing retreats reported in the literature, such as feeling empowered writers through supporting each other in a respectful and non-judgemental environment [~\cite{papen2018}], increasing our sense of belonging as PhD students [~\cite{omeara}] and decreasing our sense of isolation [~\cite{eardley2021}].
+However, most importantly, we got our writing goals for those weeks achieved.
 Some of the references herein can help with the theoretical side of writing, but ultimatelly writing is practice [~\cite{peterson2018, grogan2021}].
-By facilitating the planning and execution of writing retreats, we hope this piece can foster scholarly productivity.
+By facilitating the planning and execution of writing retreats through this Ten Simple Rules, we hope to foster scholarly productivity and wellbeing in student cohorts.
 Happy writing!
@@
 \paragraph*{Writing – original draft:}  Nicholas W.
-Daudt (lead); Claudia Hird; Eleanor Kelly; Gretchen McCarthy; Ian S.
+Daudt; Claudia Hird; Eleanor Kelly; Gretchen McCarthy; Ian S.
 Dixon-Anderson; Jackson Beagley; Joseph Curtis; Meghan Duffy; Saskia Foreman; Leah M.

### 2025-07-09 manuscript_ten-simple-rules-writing-retreat.tex
just deleted a sentence - finalised first draft
revised version of the ms, after internal round of peer review

--- 2024-08-06 manuscript_ten-simple-rules-writing-retreat.tex
@@
 {\Large
-\textbf\newline{Ten simple rules to organise an effective (student-led) writing retreat}
+\textbf\newline{Ten simple rules for organising an effective (student-led) writing retreat}
 }
@@
 \\
-Nicholas Winterle Daudt\textsuperscript{1,2*},
+Nicholas W. Daudt\textsuperscript{1,2*},
 Claudia Hird\textsuperscript{1,3},
@@
 Meghan Duffy\textsuperscript{1,4},
+Preston Maluafiti\textsuperscript{1},
 Saskia Foreman\textsuperscript{1},
+William Carome\textsuperscript{1},
 Leah M. Crowe\textsuperscript{1,2}
@@
 \bigskip
-\textbf{1} Department of Marine Science, University of Otago, Dunedin, Aotearoa New Zealand
+\textbf{1} Department of Marine Science, University of Otago, Dunedin, Aotearoa/New Zealand
 \\
-\textbf{2} Department of Mathematics and Statistics, University of Otago, Dunedin, Aotearoa New Zealand
+\textbf{2} Department of Mathematics and Statistics, University of Otago, Dunedin, Aotearoa/New Zealand
 \\
-\textbf{3} Department of Zoology, University of Otago, Dunedin, Aotearoa New Zealand
+\textbf{3} Department of Zoology, University of Otago, Dunedin, Aotearoa/New Zealand
 \\
-\textbf{4} Department of Geology, University of Otago, Dunedin, Aotearoa New Zealand
+\textbf{4} Department of Geology, University of Otago, Dunedin, Aotearoa/New Zealand
 \\
@@
 
-Writing is an integral part of academic work.
-No matter in your career stage, your scholarly output will advance scientific knowledge and count in your career development, getting grants or landing a job.
-However, in multi-tasking admin, teaching, fieldwork, among other activities, we frequently put aside how important it is to take the proper time to wrap it all up in a written document (e.g.
-thesis, scientific paper).
-Scientific writing has its technical tips and tricks, but ultimately, it is a creative act.
-Yet, the lack of time to think critically, create knowledge and transfer it into words is a significant complaint in academia [~\cite{menzies}].
-Therefore, we have to actively disengage from other tasks to enable us to properly engage with the act of writing [~\cite{murray2013}].
-In this context, writing retreats have been shown as effective opportunities to unplug from daily work routines, allowing scientists to excel in their passion: producing science.
+At every stage in a researcher's career, scholarly output advances scientific knowledge and supports career development.
+Early career researchers, in particular, significantly boost their career prospects by increasing their scholarly outputs [~\cite{horta2016, wilkins2021}].
+Writing serves as an integral skill for academic work [~\cite{boice1984, hazelett2025}], especially when competing for grants and jobs.
+Academics juggle administrative tasks alongside teaching, collection and analysis of data, and production of publications and presentations.
+Consequently, many report a lack of time to think critically as a major challenge in academia [~\cite{menzies2007}], which too often leads researchers to deprioritise writing tasks [~\cite{boice1984, menzies2007}].
+Therefore, we must actively disengage from other tasks to fully engage with the act of writing [~\cite{murray2009, murray2013}].
 
-Writing retreats are structured periods where academics have dedicated time slots for deep-focused writing [~\cite{mcgrail2006, murray2009}].
-Initially driven by the motivation to connect with our PhD cohort [~\cite{bernery, omeara}] but also wanting to leverage the moment to have a productive week of work (no pressure, but the clock does not stop ticking!), we set up for a 5-day writing retreat in a remote field station.
-Once we returned from that week, we realised how much writing we had accomplished and how important it was to share moments as PhDs going through similar experiences in various stages of our degrees.
-The motivation after that week was such that we organised a second successful retreat edition a year later.
+Writing retreats provide structured periods where academics dedicate time to focused writing [~\cite{mcgrail2006, murray2009}].
+These retreats offer practical opportunities to disconnect from daily work routines [~\cite{murray2013, tremblay2021}], which help researchers gain writing momentum and increase scholarly output [~\cite{mcgrail2006, kornhaber2016}].
+In addition, writing retreats foster a sense of community, promote wellbeing, and build self-confidence for academic writers [~\cite{kornhaber2016, eardley2021, tremblay2021}].
+Postgraduate students, in particular, highlight the value of these retreats in strengthening bonds among peers and appreciate the dedicated time and space to focus solely on writing [~\cite{kornhaber2016, papen2018, tremblay2021, bojovic2024}].
 
-Here, we build on these experiences of student-led writing retreats and lay out Ten Simple Rules for organising an effective retreat.
-Althuogh the authors are/were all PhD candidates at the time of the retreats, the rules herein should be helpful for research groups and academics more broadly.
-We cover steps to facilitate the planning/execution of pre- (Rules 1--4), during (Rules 5--9), and pos- (Rule 10) writing retreat actions, but not writing techniques per se.
-There are many resources on academic writing that we will not delve into but encourage their reading (e.g.
-[~\cite{turbek2016, hotaling2020, carter2020, cargill2021}]), including articles in this series [~\cite{weinberger2015, mensh2017, peterson2018}].
-Otherwise, you may consider writing a boring contribution instead [~\cite{sand2007}].
+As a cohort of PhD students in the Department of Marine Science at the University of Otago (Aotearoa/New Zealand), we organised a 5-day writing retreat at a remote field station in 2023.
+In New Zealand, PhDs follow research-only programs; as such, we do not participate in coursework that might promote group cohesion, and each student conducts research individually.
+In addition, given our diverse disciplines, field sites, and lab locations, we had felt a disconnection among us and a lack of community.
+Therefore, we initially planned the retreat to connect with our PhD cohort [~\cite{omeara2017, bernery2022}], but we also aimed to make it a productive work week [~\cite{mcgrail2006, kornhaber2016}].
+After returning from the retreat, we realised how much writing we had accomplished and the importance of the moments we had shared as postgrads going through similar experiences in various stages of our PhDs.
+The retreat's success motivated us to organise a second retreat the following year; the second retreat's success inspired a third.
+
+Based on our experience, we present Ten Simple Rules for organising effective (student-led) writing retreats.
+Although the authors were all PhD students at the time of the retreats, these rules apply broadly to any research-oriented or academically-minded group.
+We outline steps to support the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and post- (Rule 10) writing retreat actions, but we do not cover writing techniques per se.
+Many resources on academic writing exist that we encourage you to explore [e.g.~\cite{turbek2016, hotaling2020, carter2020, cargill2021}], including articles in this series [~\cite{weinberger2015, mensh2017, peterson2018}]---or you may simply choose to write a boring contribution instead [~\cite{sand2007}].
 
@@
 
-A change of scenery can help inspire productivity whether the writing retreat is structured as a locally organized event or as multi-day trip.
-Local facilities are often more straightforward as this can include meeting rooms and other bookable campus spaces.
-For longer retreats, students, regardless of their departmental affiliation, should inquire about their university’s field stations.
-Many biology and ecology programmes, for example, have field stations (for a global list, see [~\cite{tydecks}]), and by leveraging university owned facilities, retreat costs can be kept within a manageable budget.
-There are often times of the year when field stations are underutilized, and having another reason for people to be in a field station during these periods can help with ongoing upkeep.
-In addition, field stations are often in remote, beautiful settings amongst nature with limited distractions providing an environment conducive to focusing and connecting with peers.
-If your university doesn’t have a field station, or one that would properly support a writing retreat, other universities are often open to shared use if availability allows.
-We encourage students in this situation to inquire with station managers to build those relationships.
+A change of scenery inspires productivity [~\cite{murray2013}], whether you organise the writing retreat locally or as a multi-day trip.
+Local facilities, such as meeting rooms and other bookable spaces on campus, offer accessible options.
+For longer retreats, students---regardless of departmental affiliation---should inquire about their university's field stations.
+For example, many biology and ecology programmes maintain field stations (for a global list, see [~\cite{tydecks2016}]).
+By using university-owned facilities, students keep retreat costs manageable.
+Field stations often remain underutilized during certain periods, so retreats during these times can be used to promote year-round upkeep of facilities.
+In addition, field stations often sit in remote, natural settings with limited distractions, creating ideal environments for focus and uninterrupted work [~\cite{murray2013, tremblay2021}].
+If your university lacks a suitable field station, other institutions may allow shared use depending on availability.
+We encourage students to contact station managers and build those connections.
 
-When determining if a space is suitable for a writing retreat, it is important to take an inventory of available amenities both within the facility and in the area.
-We made use of our department’s marine field station in the town of Oban on Rakiura/Stewart Island, a small island off the southern coast of the South Island of Aotearoa New Zealand.
-This location suited us well as everything we needed was within walking distance.
-Because this is a small community (around 400 year-round residents) abutting a national forest, we had limited distractions, straightforward options for restaurants and groceries, and we had ample options for walking tracks and beach strolls.
-This field station is a three-bedroom home with twelve bunks, three bathrooms, a kitchen, and an open space with a large table, two desks, and two couches.
-Ten people attended in each of the two years which was a comfortable number for the space.
-When determining if a space will work for a retreat, it is also important to understand the internet availability, the heating/cooling systems, sleeping spaces, and work spaces (including available power points).
+When evaluating a potential site for a retreat, it is important to review available amenities both on-site and nearby.
+Also, make sure the space accommodates people with physical disabilities or impairments.
+We made use of our department's marine field station in Oban, Rakiura/Stewart Island (a small island off the southern coast of New Zealand's South Island).
+This location suited us well---everything we needed was within walking distance.
+Because this is a small community (around 400 year-round residents) abutting a national forest, we had limited distractions, straightforward options for restaurants and groceries, and ample access to walking tracks and beach strolls.
+When assessing suitability of the location, also consider internet access, heating/cooling, sleeping arrangements, and workspace infrastructure (power points, seats, tables, etc.).
+Taking advantage of institutional resources simplifies the planning process and enhances the success of your writing retreat.
 
-\section*{Rule 2: Prepare a proposal}
+\section*{Rule 2: Prepare a proposal to secure support}
 
-Before you reach out for support, it is important to be prepared and put together a proposal.
-Your proposal should include possible locations, accommodation, transport, dates and potential participants.
-Deciding how many people and specifically who the retreat is for is crucial as it is required for other planning details.
-Numbers may be decided for you based on the accommodation available (see Rule 1) or it could be the other way around.
-Once you have outlined possible locations and numbers, it is important to work out associated costs.
-Additional considerations may include transportation to, from and around the writing retreat location, food requirements, and dates/times.
+Begin planning your retreat by preparing a proposal that outlines the type of event you aim to organise and gauges interests from participants.
+Clearly define the capacity and target audience to plan details of the retreat effectively.
+Your options for accommodation (see Rule 1) may determine your maximum number of participants.
+In our case, we hosted all retreats with 10 participants; we found this number worked well for the size of our facility and the length of the retreat.
 
-Once you have written your proposal, you can reach out for support.
-The first port of call should be the departments of the students attending the retreat.
-Other bodies within the university may be able to fund and support your writing retreat such as higher divisions and postgraduate support offices.
-Industry, community organisations (e.g.
-Lions and Rotary Clubs) and fundraising may be able to provide supplemental support, if needed.
-A well-considered proposal will provide a good foundation for your retreat and set you up for success.
+After outlining potential locations, dates, and number of attendees, calculate the associated costs. These may include transportation, food, accommodation, and amenities such as internet.
+
+Use the proposal to seek support.
+Academic researchers may access departmental or divisional funding to support these opportunities for development.
+There may be external funding available to provide supplemental support if internal money is not available to cover the entire costs of the retreat.
+This funding could be sought from industry and community organisations (e.g.
+Lions and Rotary Clubs), or through fundraisers.
+A well-developed proposal lays a solid foundation for organising the retreat and securing the support needed for success.
 
@@
 
-We were fortunate that our department was very supportive of the writing retreat proposal, and when planning the first trip, encouraged us to use experts on staff at the university.
-At the University of Otago, we worked with the Higher Education Development Centre which provided resources and guidance on how to structure both productive days and the overall week.
-There are many resources available to guide writing retreats for PhD students, including this one, but we encourage student to take advantage of these kinds of services at their universities if they are available.
+Create a schedule to structure your writing retreat, as this will be key to its success [~\cite{murray2009, tremblay2021}].
+Use available resources to guide your planning [e.g.
+~\cite{tremblay2021}], including this guide (see also Appendices).
+At our university, we collaborated with the Higher Education Development Centre, which provided guidance on designing productive days and structuring the week for our first and second retreats.
+We encourage you to take advantage of such services when available.
 
-Our week was organized so that we traveled to and from the field station on Sunday and Saturday to allow for Monday through Friday to be fully structured for the retreat.
-We left extra time on travel days to settle in/pack up, grocery shop, and explore the area.
-Upon arrival, it was important to have a short welcome discussion to establish ground rules and make sure everyone was comfortable in the space (see Rule 5).
-Each work day was themed to give some guidance on what to try and tackle at that point in the week (Table~\ref{table1}).
-We also made sure to intersperse optional fun activities throughout the week which included a pub quiz (i.e.
-bar trivia), wildlife tours, game night, and other local events.
+We travelled to the accommodation on Sunday and returned the following Saturday, allowing Monday through Friday to remain fully structured for retreat activities.
+We built in extra time on travel days to settle in, pack up, shop for groceries, and explore the area.
+Upon arrival, we held a welcome discussion to establish ground rules and ensure everyone felt comfortable (see Rule 5).
+We themed each workday to provide guidance and structure for participants (Table~\ref{table1}), while allowing each individual to pursue their own goals.
+We also incorporated optional social activities throughout the week (see Rule 9), such as a pub quiz, wildlife tours, game night, and local events.
 
@@
 \centering
-\caption{
-{\bf Daily themes over a 5-day writing retreat.}}
+\caption{{\bf Examples of daily themes over a 5-day writing retreat. Note that days 1 and 7 were organised for travel to and settling in, and packing up and travelling back, respectively.}}
 \begin{tabular}{p{1in}p{2in}p{2in}}
 \hline
-{\bf Weekday} & {\bf Theme} & {\bf Description}\\ \thickhline
-Monday & Getting started & Commit to the goal for the week (see Rule 4), organize a strategy, and establish accountability partners.\\ \hline
-Tuesday & Words on the page & Just start writing something! (Rule 6)\\ \hline
-Wednesday & Keep going! & Continue writing and refine. (Rule 6)\\ \hline
-Thursday & Good enough & Exchange work within accountability partners to practice giving and receiving comments.
-Use this opportunity to reflect on writing structure and style from a reader’s perspective.
+{\bf Day} & {\bf Theme} & {\bf Description}\\ \thickhline
+Day 2 & Getting started & Commit to the goal for the week (see Rule 4), organize a strategy, and establish accountability partners.\\ \hline
+Day 3 & Words on the page & Just start writing something! (Rule 6)\\ \hline
+Day 4 & Keep going! & Continue writing and refine. (Rule 6)\\ \hline
+Day 5 & Good enough & Exchange work within accountability partners to practice giving and receiving comments.
+Use this opportunity to reflect on writing structure and style from a reader's perspective.
 (Rule 8)\\ \hline
-Friday & Wrap it up & End the retreat by reflecting on what was accomplished this week (Rule 10) and develop a plan on how to proceed.\\ \hline
+Day 6 & Wrap it up & End the retreat by reflecting on what was accomplished this week (Rule 10) and develop a plan on how to proceed.\\ \hline
 \end{tabular}
@@
 
-The schedule for each day was roughly between 9 am and 6 pm and was provided in advance of the retreat so participants knew what to expect for dedicated work time (see \nameref{S1_Appendix}).
-A different leader was designated for each day to lead discussions and keep the time for writing blocks (between 60 and 120 min) and breaks (between 30 and 60 min).
-We started each morning together by writing for 15 minutes using prompts that varied from reflections on the different ways we write in our lives to the most ridiculous title we could conjure for our research.
-This was a productive way to get our writing engine revving and a fun opportunity to share writing with each other.
-Each evening we ended the day by a reflective discussion on how the day went (with an 'accountability buddy' [e.g.
-'Rule 5' in ~\cite{peterson2018}]), a shared dinner, and presentations (see Rule 7).
+Our daily schedule typically ran from 9:00 AM to 6:00 PM and was shared in advance so participants knew when to expect focused work sessions [~\cite{tremblay2021}; see also \nameref{S1_Appendix}].
+We designated a different facilitator each day to lead discussions and manage writing blocks (60--120 minutes) and breaks (30--60 minutes).
+Each morning began with a 15-minute writing session using prompts that ranged from reflective writing to humorous research titles (see Rule 9).
+These sessions helped rev up our writing engines for the day and offered a fun, low-pressure opportunity to share writing.
+Each evening, we closed the day with a reflective discussion on how the day went (with an `accountability partner' [see Rule 5 in ~\cite{peterson2018}]), a shared dinner, and presentations (see Rule 7).
 
@@
 
-Hosting a pre-retreat meeting with attendees before departure facilitates a smoother and more productive retreat by addressing the primary objectives of introductions, finalizing logistics, and goal-setting.
-During the event, participants will get a chance to meet (if they haven’t already) and establish or reinforce a familiar group dynamic, an essential prerequisite for a retreat atmosphere of respect and accountability (see Rules 5 and 10).
-Fine-tuning the plan for retreat logistics and scheduling through group conversation can identify opportunities for clarification and improvement, and create a sense of shared ownership over day-to-day programming.
-Most importantly, a pre-retreat meeting provides a chance for participants to identify or workshop specific writing goals with time to prepare relevant resources (literature, data analysis, input of academic supervisors) so that these non-writing activities don't become distractions during the initial days of the retreat.
+Host a pre-retreat meeting with all attendees to ensure a smoother and more productive retreat.
+This meeting gives participants a chance to meet, finalise logistics, set goals, and raise questions or concerns.
+Use this time to build or reinforce a respectful and accountable group dynamic (see Rules 5 and 10).
+Collaboratively fine-tune logistics and schedules to clarify plans and foster shared ownership.
+Most importantly, a pre-retreat meeting allows participants to identify or workshop specific writing goals with time to prepare relevant resources (literature, data analysis, input from academic supervisors) so that non-writing activities do not become distractions during the initial days of the retreat.
 
-We suggest participants should arrive at the pre-retreat meeting with a preliminary goal to share with the group.
-Goals can be quantitative (word count, page number) or qualitative (paper sections, revision of existing work).
-For reference, the average output during our 5-day retreats was 1.2 pages or 550 words per day, and ranged from refined work on a single chapter/manuscript section to drafts of entire manuscripts.
-Critically, the specific thesis component (or other written work) should be identified at the meeting to remove the friction of choosing a piece of writing at the outset of the retreat and maximizing preparation time.
-Sharing specific goals has the additional benefits of introducing the group to everybody's topic of research, as well as identifying whether a participant's preliminary goal is appropriately ambitious for an extended period of dedicated writing.
+We suggest that participants arrive at the pre-retreat meeting with a preliminary goal to share with the group.
+Goals may be quantitative (e.g.
+word count, page number) or qualitative (e.g.
+specific sections or revisions).
+For reference, during our 5-day retreats, participants averaged 1.2 pages or 550 words per day, working on everything from detailed revisions to drafting full manuscripts.
+Identifying a specific writing target ahead of time reduces decision-making friction at the retreat and maximises productive preparation.
+Sharing goals also helps participants learn about each other's research and assess whether goals are realistic.
+Hold the pre-retreat meeting within two weeks of departure.
+This timeframe supports goal-oriented preparation while preserving participant momentum.
+If possible, host the meeting in person in a casual setting.
+If attendees join from different campuses, a video call works as well.
+Additional objectives may include selecting writing prompts, presentation topics, icebreakers, roles (e.g.
+meal coordination, daily facilitator), or ideas for social activities.
 
-We recommend that a pre-retreat meeting be held no more than two weeks before departure, allowing time for goal-oriented preparation while limiting stagnation of attention and momentum heading into the retreat.
-If possible, hosting the event in person (ideally off-campus) is preferrable, although most of the aims of the meeting could be achieved via video conference if necessary (for instance, if you have people from other campi attending the retreat).
-Secondary objectives for a pre-retreat meeting could include selecting creative writing prompts, topics for presentations (if these are included in your retreat), "get to know you" activities, signing up for roles (meal spots, daily facilitator), or crowd-sourcing ideas for social aspects such as field trips or evening activities.
-If a planning event cannot be arranged before the retreat, objectives of the meeting (introductions, goal-setting, flexible logistic planning) should still be performed during the first day of the retreat (likely just after discussing the ground rules; see Rule 5), though this will come at the expense of writing time, momentum, and opportunities for preparation.
+If you cannot host a pre-retreat meeting, incorporate these objectives (introductions, goal-setting, and planning) into the first day of the retreat, ideally following the discussion of ground rules (see Rule 5).
+However, doing so may reduce writing time and diminish early momentum.
 
@@
 
-At the start of your retreat, it is important to establish clear ground rules and expectations.
-These can include shared expectations for the group, such as setting quiet hours, assigning household responsibilities (i.e.
-cooking and cleaning) and setting guidelines for breaks.
-The document used as the basis for our ground rules can be found in \nameref{S2_Appendix}.
-Ground rules should also set the tone of the retreat, creating a balanced foundation for respect, productivity and enthusiasm.
+Establish clear ground rules and expectations at the start of the retreat.
+These include shared expectations for the group, such as setting quiet hours, assigning household responsibilities (e.g.
+cooking and cleaning), and setting guidelines for breaks.
+You can find the document we used as the basis for our ground rules in \nameref{S2_Appendix}.
+Ground rules should also set the tone of the retreat, creating a balanced foundation for respect, productivity, and enthusiasm.
 
-At our retreat, the primary guideline we established was a focus on respectful and non-judgemental interactions between all participants.
+At our retreat, we prioritised respectful and non-judgemental interactions between all participants.
 This attitude allowed flexibility in goal setting and accomplishments depending on each person's needs and stage of their PhD.
-The baseline expectation of respect also led to creating clear expectations for working versus quiet hours and maintaining a productive working environment.
-During our ground rule meeting, we also discussed the need for flexibility to ensure our retreat catered to the needs of all the participants.
-For example, not all participants are native English speakers and may find full days of writing and discussion in English more tiring.
-As a result, we designated certain writing blocks as optional---if participants needed a longer break or felt their time would be better spent refreshing their mind with a forest walk or gym session, there was no judgement.
-Continuing with the guideline of respect, our opening meeting also made it clear that while group discussions and feedback are useful, only constructive criticism would be tolerated.
-Together, clear expectations, a set schedule and overall attitude of respect set the tone for a successful week of writing.
+Our baseline expectation of respect allowed us to set clear expectations for working versus quiet hours and to maintain a productive and peaceful environment.
+During our ground rule meeting, we also discussed the need for flexibility to ensure our retreat catered to the needs of all participants.
+For example, not all participants were native English speakers and may find full days of writing and discussion in English more tiring.
+As a result, we designated writing blocks as optional---if participants need a longer break or feel they would benefit from refreshing their mind with a forest walk or gym session, they should do so without judgement.
+Continuing with our guideline of respect, we also made it clear in our opening meeting that while group discussions and feedback are useful, we only tolerate constructive criticism.
+Together, clear expectations, a set schedule, and an overall attitude of respect set the tone for a successful week of writing.
 
@@
 
-This is why you wanted to organise a retreat in the first place---to make progress on your academic writing.
-You have already created a schedule and a daily theme for yourself (see Rule 3) so it’s time to follow it.
-This is where the goals that you have set for yourself (see Rule 4) also come into play.
-Knowing what you are working towards, and having a plan of how you intend to get there, should provide you with the space needed to put words on the page and have productive writing blocks.
-However, writing is not easy and writers face many hurdles[~\cite{grogan2021}], such as a lack of motivation, uncertainty about what to do next, writer’s block, imposter syndrome, and more.
-These obstacles can easily disrupt even the best-laid plans.
-It is therefore important to have a backup plan.
+Write.
+This is why you organise a retreat in the first place---to make progress on academic writing.
+Once a daily schedule and theme are set (Rule 3), it is time to execute.
+This is where individual goals (Rule 4) come into play.
+Knowing what you are working towards, and having a plan to get there, provides the bandwidth needed to put words on the page and have productive writing blocks.
 
-If you are struggling to get any writing done it may be a good time to take a break or switch tasks.
-It is proven that even short breaks can increase productivity [~\cite{lyubykh, carter2020}].
-Alternatively, this could be a good time for you to do some research into writing development.
-If you are not being productive in writing, you may as well be productive in learning how to write.
-Everybody has different strategies that work for them.
-For example, some people prefer to cite as they write, whereas others prefer to write everything out quickly and then add citations as they edit.
-Some authors, such as Ernest Hemingway [~\cite{hemingway}], have even claimed that they purposely leave a sentence, paragraph, or idea unfinished at the end of their writing day specifically so they have something easy to start with for their next writing block.
-It is important to understand what works for you in terms of a writing practice and to reevaluate your process often to ensure it is still working [~\cite{peterson2018, grogan2021}].
+However, writing is not easy, and writers face many hurdles, such as a lack of motivation, uncertainty about what to do next, writer's block, imposter syndrome, and more [~\cite{boice1984, murray2013, wilson2019, grogan2021}].
+These obstacles can easily disrupt even the best-laid plans and most robust retreat structures.
+Therefore, have a backup plan.
+If it becomes challenging to write, consider taking a break or switching tasks; it is proven that even short breaks increase productivity [~\cite{carter2020, lyubykh2022}].
+Everyone has different strategies that work for them.
+For example, some people prefer to cite as they write, while others write everything out quickly and add citations during editing.
+Some authors, such as Ernest Hemingway, have even suggested to leave a sentence, paragraph, or idea unfinished at the end of a writing day so they have something easy to start with during the next writing block [~\cite{hemingway1935}].
+Assess what works best for your writing practice and reevaluate regularly to ensure that strategy is still serving you [~\cite{peterson2018, grogan2021}].
 
@@
 
-While the primary objective of a writing retreat is to write, there are also opportunities for other forms of professional development.
-Undertaking the PhD journey is daunting and different for every student, and it is easy to become so engrossed in your own work, that you forgot you are surrounded by people in the same situation as you.
+Schedule academic discussions around participants' work in a structured manner.
+A writing retreat gathers your academic peers in a collaborative environment with minimal external distractions.
+It presents a unique opportunity to garner advice and perspectives on written work (see also Rule 8) or any work-in-progress that participants might be developing.
+
+Undertaking any large-writing project (e.g.
+a PhD thesis) is a daunting and varying experience for everyone.
+It is easy to become so engrossed in your work that you forget you are surrounded by people in a similar situation [~\cite{wilson2019}].
 A writing retreat provides the opportunity to share your own research or discuss scientific/academic topics amongst your peers.
-While we found this occurs naturally throughout the retreat, especially within groups of similarly focused researchers, it was decided that a small block of time each day (see Rule 3) would be specifically allocated for students to share a talk or lead a discussion of their own with the other attendees.
-The casual, non-judgemental environment gave presenting students a chance to practice a seminar talk, get insight on methodology, analysis and results, or simply share "tips and tools of the trade".
-For many students, this was an opportunity to learn about the research their peers conduct; in a field as multi-faceted as marine science, it is common that we are unaware of the scope or type of research our peers are involved in.
+While we found this occurred naturally throughout the retreats, especially within groups of similarly focused researchers, we also allocated a small block of time each day (see Rule 3) specifically for students to share a talk or lead a discussion with the other attendees.
+The casual, non-judgemental environment allowed presenting students a chance to practice a seminar talk, gain insight on methodologies, analyses and results, or simply share `tips and tools of the trade.'
 
-We suggest scheduling the discussions after writing blocks are complete for the day and spread them out over the retreat.
-Access to a projector and screen in our second retreat was beneficial.
-However, previously, we have simply shared from our laptop screens or consulted notes and found it just as engaging due to the small group attending the retreat.
-Talks were not strictly timed as it allowed for conversation throughout, but we aimed not to exceed 20 min per discussion.
+We suggest scheduling these discussions after writing blocks are complete for the day and spreading them out over the retreat.
+Access to a projector and screen in our second and third retreats was beneficial.
+However, previously, we have simply shared from our laptop screens or consulted notes and found it equally engaging due to the small group attending the retreats.
+Discussions were informal, and we aimed not to exceed 20 minutes per session.
 It is important to emphasise the casualness of such an exercise---students should not be using their time for writing anxiously preparing a conference-level presentation.
-It is simply an opportunity to share, learn, provide and gain insight.
-Importantly, these sections give a chance get to know one another and bound as fellow PhD students [~\cite{bernery, omeara}].
+These are simply opportunities to share, learn, and exchange insights.
+Importantly, these sessions offer a chance to get to know one another and connect as fellow PhD students [~\cite{omeara2017, bernery2022}].
 
@@
 
-An important aspect of writing is the practice of giving and receiving revisions with your peers.
-It is important to remember that the feedback exchanged between peers should be constructive and can offer insight on gaps in clarity, logic, or structure.
-The use of neutral language should be encouraged [~\cite{parsons2021}].
-There is the possibility you may not understand content your peer is working on, however, focusing on aspects of structure, prose, and overall writing flow is a good place to start with peer review and revisions.
-Receiving feedback from a peer outside of their field may provide different perspectives the author had not previously considered and can lead to more balanced and comprehensive writing.
-It is also important to receive feedback as much as you can provide it as this practice will encourage a collaborative approach to writing which can build confidence and resilience in both the reviewed and the reviewer.
+Schedule time to exchange written material and review that of your peers.
+An essential aspect of developing your writing is giving and receiving revisions [~\cite{guilford2001, reynolds2011}].
+Constructive feedback from peers may offer valuable insights on gaps in clarity, logic, and structure.
+Even if you do not understand the scientific content of the work, focusing on structure, prose, and writing flow can be extremely helpful.
+This type of review can unblock a colleague who feels stuck.
+Likewise, feedback from a peer outside one's field may offer perspectives the author had not previously considered.
+
+Providing feedback is just as valuable as receiving it.
+This practice encourages a collaborative approach to writing that builds confidence and resilience in both the reviewer and the reviewed.
+When giving feedback, use neutral language [~\cite{parsons2021}] and focus on writing components (structure, clarity, flow).
+If you are an expert in the content, detailed feedback can also be extremely valuable.
+Academics in general should be mindful of the historical barriers peer review imposes on minority groups and non-English speakers [~\cite{smith2023, saleh2024}], and work towards breaking this pattern.
+Furthermore, peer reviewing should be an integral part of academic life [~\cite{borja2024}].
+Recognise peer review as a skill that can be practised, and a writing retreat could be the perfect place to begin if you are new to it.
 
@@
 
-A key aspect of a writing retreat is to have fun and enjoy the whole experience---from writing to exploring the local area.
-While on a writing retreat, it is important to take opportunities to experience new activities you may not usually get to.
-A retreat also presents a unique chance to bond with your peers while balancing downtime activities and writing time for optimal productivity.
+Have fun and build connections with your peers.
+While on a writing retreat, including recreational activities can enhance the overall experience.
+A writing retreat allows you to disconnect from the routine and responsibilities [~\cite{murray2013}], providing a unique opportunity to bond with your peers.
+More rural places often have scenic landscapes, hospitable communities and limited distractions.
+While this may restrict choices for activities, it can lead to higher quality experiences where you can get to know local businesses and each other on a more personal level.
+Before embarking on a writing retreat, research what is unique about the location and take advantage of what may be on offer.
+We organised creative writing prompts for each morning (see Rule 3) and gave casual presentations (Rule 7) about various topics, which were fun and engaging bonding activities.
+Furthermore, balancing downtime activities with writing blocks can lead to optimal productivity [~\cite{eardley2021, lyubykh2022}], enhancing the writing experience.
 
-Before you embark on your writing retreat, we recommend that you do some research before arriving to see what is unique about the area to take advantage of what may be on offer.
-While based at our university’s field station, we had opportunities to aurora hunt and kiwi spot, explore the nearby Ulva island (a protected area), participate in bar trivia, and develop relationships with some of the locals.
-We also organised creative morning writing prompts (see Rule 3) and gave casual presentations (Rule 7) about various topics which were fun and engaging bonding activities.
+\section*{Rule 10: Posterity—gather feedback and demonstrate value}
 
-More rural places often have scenic landscapes, hospitable communities and limited distractions.
-This could restrict activity choices but may produce more quality experiences where you can get to know the local businesses and people on a more personal level.
-A retreat is not only about writing but an opportunity to engage with peers and have fun.
+After the retreat, take time to reflect and gather feedback from participants.
+A group discussion about the retreat's highlights and challenges may help clarify thoughts before participants are asked to complete an anonymous survey.
+Shortly after returning, summarise the feedback into a concise report while details are still fresh.
+Be sure to include comments on the schedule and notes on things that went well or poorly.
+This is valuable for cementing ideas and details that would otherwise fade before another retreat can be planned.
+These insights will shape future retreats and ensure important lessons are not lost over time.
 
-\section*{Rule 10: Posterity -- Be accountable and acknowledge}
+A report also creates accountability and continuity.
+Student organisers may change, but a shared archive of resources---such as schedules, ground rules, and facilitation tips---makes it easier to plan future events.
+Creating an institutional repository for retreat materials provides blueprints and templates to future organisers, setting them up for success.
 
-Gathering feedback from participants of the retreat and drafting a report are important actions following the writing retreat.
-Allowing time for the participants to reflect on and discuss to their peers the pitfalls and peaks of the retreat on the last day or two of the retreat and then having them fill out a anonymous survey immediately after returning is very beneficial.
-By summarizing these points into a concise report soon after attending the retreat, points are clear in the minds of the participants and key issues or takeaways are not forgotten.
-Comments of the schedule, notes on things that went well, and notes of things that might need improvements are valuable to cement ideas and details that will fade in time before another retreat can be planned.
-This will reinforce great points from the retreat but more importantly creates some accountability for posterity.
-Knowledge transfer from one retreat to the next is difficult in an ever-changing student body, but creating a paper trail of advice and resources within your own department provides blueprints and templates for new volunteers to organize their own retreats.
-%The post-retreat report can act as a manual for future students and helps to lower the barrier to organizing a productive event by being a reference to future leaders with answers to FAQs and examples of how to organize administrative tasks.
-
-In addition, give acknowledgement to the writing retreat in your work and publications and circulate this knowledge to your peers, your department, and your colleagues.
-Having acknowledgements of the retreat (potentially with a uniform phrase) can even provide a searchable tool in online databases.
-Creating a repository where work that has benefited from the writing retreat can be achieved by filtering these acknowledgements and compiling publications together in the same place.
-Highlighting to your department the quantity and quality of work that results from a writing retreat is key to securing the longevity of writing retreats.
-A list which updates into the future as new work is published that stemmed from the writing retreat means that administrative personnel can become aware of the significant impact this opportunity has provided student researchers.
-Both the scientific community and the university administration often use publication numbers as their tangible metric to interpret how their financial decisions have made an impact.
-Evidence of the how many papers may come out of a writing retreat by acknowledging the usefulness of it will help to capitalizing on this trend.
-A scientifically-productive retreat may ease the justification process of future student organizers, or even incentivise greater financial support from administrative groups if the expenditure of writing retreats are seen as a net-positive for publications.
+In addition, publicly acknowledging the retreat in your writing and presentations demonstrates its academic value.
+Include it in the acknowledgements of manuscripts and theses worked on during the retreat.
+Sharing this impact helps advocate for continued support, especially in institutions where publication output informs resource allocation.
+Maintaining a list of retreat-supported outputs can strengthen future proposals (see Rule 2) and make a compelling case for funding.
+Indeed, demonstrating the immense value of this experience may spur on support from leadership to fund retreats for students in other departments or institutions.
 
@@
 
-Following these set of rules, our writing retreats were sucessful.
-Colectivelly, we indeed experienced benefits reported in the literature, such as feeling empowered writers through supporting each other in a respectful and non-judgemental environment [~\cite{papen2018}], increasing our sense of belonging as PhD students [~\cite{omeara}] and decreasing our sense of isolation [~\cite{eardley2021}].
-However, most importantly, we got our writing reatreat goals achieved.
-Some of the references herein can help with the theoretical side of writing, but ultimatelly writing is practice [~\cite{peterson2018, grogan2021}].
-By facilitating the planning and execution of writing retreats, we hope this piece can foster scholarly productivity.
+Collectively, we experienced myriad benefits of writing retreats reported in the literature.
+These include feeling empowered as writers through supporting each other in a respectful and non-judgmental environment [~\cite{papen2018}], increasing our sense of belonging as PhD students [~\cite{omeara2017}] and decreasing our sense of isolation [~\cite{eardley2021}].
+Most importantly, we achieved our writing goals at the retreats.
+Some of the references herein can help with the theoretical side of writing, but ultimately writing is practice and requires dedicated time [~\cite{peterson2018, grogan2021}].
+By facilitating the planning and execution of writing retreats through these Ten Simple Rules, we hope to foster scholarly productivity and wellbeing in student cohorts.
 Happy writing!
@@
 \label{S1_Appendix}
-{\bf Schedule Example.} Brief description...
+{\bf Schedule Examples.} Daily schedules used during our writing retreats in 2023, 2024 and 2025.
 
@@
 \label{S2_Appendix}
-{\bf Ground Rules.} Brief description...
+{\bf Ground Rules.} The base document for the Ground Rules discussed at the start of the retreats. Any additional points can be added to tailor the specifics of each group/location.
 
 \section*{Acknowledgments}
-This manuscript is a contribution conceptualized, led, and written by PhD students, who attended writing retreats supported by the Department of Marine Science of the University of Otago, Aotearoa New Zealand.
-We hope writing retreats stay as a trend, fostering collaboration and friendship among students and promoting quality science.
-Thanks also to Tracy Rogers and Claire Hudson from the Higher Education Development Centre of the University of Otago for the support while organising both writing retreats, and the Southern Sea Hotel and the Rakiura/Stewart Island community for welcoming us.
+This manuscript is a contribution conceptualised, led, and written by PhD students, who attended writing retreats supported by the Department of Marine Science of the University of Otago, Aotearoa/New Zealand.
+We hope writing retreats stay as a trend in our Department, fostering collaboration and friendship among students and promoting quality science.
+Thanks also to Tracy Rogers and Claire Hudson from the Higher Education Development Centre of the University of Otago for the support while organising the first and second writing retreats, and the South Sea Hotel and the Rakiura/Stewart Island community for welcoming us.
 
@@
 
-\paragraph*{Conceptualization:} Nicholas W.
-Daudt (lead); Claudia Hird; Eleanor Kelly; Elli Leinikki; Gretchen McCarthy; Ian S.
+\paragraph*{Conceptualisation:} Nicholas W.
+Daudt (lead); Claudia Hird; Eleanor Kelly; Gretchen McCarthy; Ian S.
 Dixon-Anderson; Jackson Beagley; Jessica Moffitt; Joseph Curtis; Lindsay Wickman; Meghan Duffy; Saskia Foreman; Leah M.
@@
 
-\paragraph*{Project administration:} Nicholas W. Daudt.
-
 \paragraph*{Writing – original draft:} Nicholas W.
-Daudt (lead); Claudia Hird; Eleanor Kelly; Gretchen McCarthy; Ian S.
+Daudt; Claudia Hird; Eleanor Kelly; Gretchen McCarthy; Ian S.
 Dixon-Anderson; Jackson Beagley; Joseph Curtis; Meghan Duffy; Saskia Foreman; Leah M.
@@
 Daudt (lead); Claudia Hird; Eleanor Kelly; Elli Leinikki; Gretchen McCarthy; Ian S.
-Dixon-Anderson; Jackson Beagley; Jessica Moffitt; Joseph Curtis; Lindsay Wickman; Meghan Duffy; Saskia Foreman; Leah M.
+Dixon-Anderson; Jackson Beagley; Jessica Moffitt; Joseph Curtis; Lindsay Wickman; Meghan Duffy; Preston Maluafiti; Saskia Foreman; William Carome; Leah M.
 Crowe.
+
+\paragraph*{Project administration:} Nicholas W. Daudt.
 
@@
 % step-by-step instructions.
+
+%\begin{thebibliography}{10}
 %
-%\begin{thebibliography}{10}
-
 %\bibitem{tydecks}
@@
 %\newblock BioScience 2016;66(2):164--171.
-
+%
 %\bibitem{hemingway}
@@
 %\newblock Esquire, Oct; 1935.
-
+%
 %\bibitem{lyubykh}
@@
 %\newblock J. Occup. Health Psychol. 2022;27(5):470--487.
-
+%
 %\end{thebibliography}

### 2025-07-10 manuscript_ten-simple-rules-writing-retreat.tex
removed double brackets around citations

@@
 At every stage in a researcher's career, scholarly output advances scientific knowledge and supports career development.
-Early career researchers, in particular, significantly boost their career prospects by increasing their scholarly outputs [~\cite{horta2016, wilkins2021}].
-Writing serves as an integral skill for academic work [~\cite{boice1984, hazelett2025}], especially when competing for grants and jobs.
+Early career researchers, in particular, significantly boost their career prospects by increasing their scholarly outputs ~\cite{horta2016, wilkins2021}.
+Writing serves as an integral skill for academic work ~\cite{boice1984, hazelett2025}, especially when competing for grants and jobs.
 Academics juggle administrative tasks alongside teaching, collection and analysis of data, and production of publications and presentations.
-Consequently, many report a lack of time to think critically as a major challenge in academia [~\cite{menzies2007}], which too often leads researchers to deprioritise writing tasks [~\cite{boice1984, menzies2007}].
-Therefore, we must actively disengage from other tasks to fully engage with the act of writing [~\cite{murray2009, murray2013}].
+Consequently, many report a lack of time to think critically as a major challenge in academia ~\cite{menzies2007}, which too often leads researchers to deprioritise writing tasks ~\cite{boice1984, menzies2007}.
+Therefore, we must actively disengage from other tasks to fully engage with the act of writing ~\cite{murray2009, murray2013}.
 
-Writing retreats provide structured periods where academics dedicate time to focused writing [~\cite{mcgrail2006, murray2009}].
-These retreats offer practical opportunities to disconnect from daily work routines [~\cite{murray2013, tremblay2021}], which help researchers gain writing momentum and increase scholarly output [~\cite{mcgrail2006, kornhaber2016}].
-In addition, writing retreats foster a sense of community, promote wellbeing, and build self-confidence for academic writers [~\cite{kornhaber2016, eardley2021, tremblay2021}].
-Postgraduate students, in particular, highlight the value of these retreats in strengthening bonds among peers and appreciate the dedicated time and space to focus solely on writing [~\cite{kornhaber2016, papen2018, tremblay2021, bojovic2024}].
+Writing retreats provide structured periods where academics dedicate time to focused writing ~\cite{mcgrail2006, murray2009}.
+These retreats offer practical opportunities to disconnect from daily work routines ~\cite{murray2013, tremblay2021}, which help researchers gain writing momentum and increase scholarly output ~\cite{mcgrail2006, kornhaber2016}.
+In addition, writing retreats foster a sense of community, promote wellbeing, and build self-confidence for academic writers ~\cite{kornhaber2016, eardley2021, tremblay2021}.
+Postgraduate students, in particular, highlight the value of these retreats in strengthening bonds among peers and appreciate the dedicated time and space to focus solely on writing ~\cite{kornhaber2016, papen2018, tremblay2021, bojovic2024}.
 
@@
 In addition, given our diverse disciplines, field sites, and lab locations, we had felt a disconnection among us and a lack of community.
-Therefore, we initially planned the retreat to connect with our PhD cohort [~\cite{omeara2017, bernery2022}], but we also aimed to make it a productive work week [~\cite{mcgrail2006, kornhaber2016}].
+Therefore, we initially planned the retreat to connect with our PhD cohort ~\cite{omeara2017, bernery2022}, but we also aimed to make it a productive work week ~\cite{mcgrail2006, kornhaber2016}.
 After returning from the retreat, we realised how much writing we had accomplished and the importance of the moments we had shared as postgrads going through similar experiences in various stages of our PhDs.
@@
 We outline steps to support the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and post- (Rule 10) writing retreat actions, but we do not cover writing techniques per se.
-Many resources on academic writing exist that we encourage you to explore [e.g.~\cite{turbek2016, hotaling2020, carter2020, cargill2021}], including articles in this series [~\cite{weinberger2015, mensh2017, peterson2018}]---or you may simply choose to write a boring contribution instead [~\cite{sand2007}].
+Many resources on academic writing exist that we encourage you to explore (e.g.
+~\cite{turbek2016, hotaling2020, carter2020, cargill2021}), including articles in this series ~\cite{weinberger2015, mensh2017, peterson2018}---or you may simply choose to write a boring contribution instead ~\cite{sand2007}.
 
@@
 
-A change of scenery inspires productivity [~\cite{murray2013}], whether you organise the writing retreat locally or as a multi-day trip.
+A change of scenery inspires productivity ~\cite{murray2013}, whether you organise the writing retreat locally or as a multi-day trip.
 Local facilities, such as meeting rooms and other bookable spaces on campus, offer accessible options.
 For longer retreats, students---regardless of departmental affiliation---should inquire about their university's field stations.
-For example, many biology and ecology programmes maintain field stations (for a global list, see [~\cite{tydecks2016}]).
+For example, many biology and ecology programmes maintain field stations (for a global list, see ~\cite{tydecks2016}).
 By using university-owned facilities, students keep retreat costs manageable.
 Field stations often remain underutilized during certain periods, so retreats during these times can be used to promote year-round upkeep of facilities.
-In addition, field stations often sit in remote, natural settings with limited distractions, creating ideal environments for focus and uninterrupted work [~\cite{murray2013, tremblay2021}].
+In addition, field stations often sit in remote, natural settings with limited distractions, creating ideal environments for focus and uninterrupted work ~\cite{murray2013, tremblay2021}.
 If your university lacks a suitable field station, other institutions may allow shared use depending on availability.
@@
 
-Create a schedule to structure your writing retreat, as this will be key to its success [~\cite{murray2009, tremblay2021}].
-Use available resources to guide your planning [e.g.
-~\cite{tremblay2021}], including this guide (see also Appendices).
+Create a schedule to structure your writing retreat, as this will be key to its success ~\cite{murray2009, tremblay2021}.
+Use available resources to guide your planning (e.g.
+~\cite{tremblay2021}), including this guide (see also Appendices).
 At our university, we collaborated with the Higher Education Development Centre, which provided guidance on designing productive days and structuring the week for our first and second retreats.
@@
 
-Our daily schedule typically ran from 9:00 AM to 6:00 PM and was shared in advance so participants knew when to expect focused work sessions [~\cite{tremblay2021}; see also \nameref{S1_Appendix}].
+Our daily schedule typically ran from 9:00 AM to 6:00 PM and was shared in advance so participants knew when to expect focused work sessions ~\cite{tremblay2021} (see also \nameref{S1_Appendix}).
 We designated a different facilitator each day to lead discussions and manage writing blocks (60--120 minutes) and breaks (30--60 minutes).
@@
 These sessions helped rev up our writing engines for the day and offered a fun, low-pressure opportunity to share writing.
-Each evening, we closed the day with a reflective discussion on how the day went (with an `accountability partner' [see Rule 5 in ~\cite{peterson2018}]), a shared dinner, and presentations (see Rule 7).
+Each evening, we closed the day with a reflective discussion on how the day went (with an `accountability partner'; see Rule 5 in ~\cite{peterson2018}), a shared dinner, and presentations (see Rule 7).
 
@@
 Sharing goals also helps participants learn about each other's research and assess whether goals are realistic.
+
 Hold the pre-retreat meeting within two weeks of departure.
@@
 
-However, writing is not easy, and writers face many hurdles, such as a lack of motivation, uncertainty about what to do next, writer's block, imposter syndrome, and more [~\cite{boice1984, murray2013, wilson2019, grogan2021}].
+However, writing is not easy, and writers face many hurdles, such as a lack of motivation, uncertainty about what to do next, writer's block, imposter syndrome, and more ~\cite{boice1984, murray2013, wilson2019, grogan2021}.
 These obstacles can easily disrupt even the best-laid plans and most robust retreat structures.
 Therefore, have a backup plan.
-If it becomes challenging to write, consider taking a break or switching tasks; it is proven that even short breaks increase productivity [~\cite{carter2020, lyubykh2022}].
+If it becomes challenging to write, consider taking a break or switching tasks; it is proven that even short breaks increase productivity ~\cite{carter2020, lyubykh2022}.
 Everyone has different strategies that work for them.
 For example, some people prefer to cite as they write, while others write everything out quickly and add citations during editing.
-Some authors, such as Ernest Hemingway, have even suggested to leave a sentence, paragraph, or idea unfinished at the end of a writing day so they have something easy to start with during the next writing block [~\cite{hemingway1935}].
-Assess what works best for your writing practice and reevaluate regularly to ensure that strategy is still serving you [~\cite{peterson2018, grogan2021}].
+Some authors, such as Ernest Hemingway, have even suggested to leave a sentence, paragraph, or idea unfinished at the end of a writing day so they have something easy to start with during the next writing block ~\cite{hemingway1935}.
+Assess what works best for your writing practice and reevaluate regularly to ensure that strategy is still serving you ~\cite{peterson2018, grogan2021}.
 
@@
 a PhD thesis) is a daunting and varying experience for everyone.
-It is easy to become so engrossed in your work that you forget you are surrounded by people in a similar situation [~\cite{wilson2019}].
+It is easy to become so engrossed in your work that you forget you are surrounded by people in a similar situation ~\cite{wilson2019}.
 A writing retreat provides the opportunity to share your own research or discuss scientific/academic topics amongst your peers.
@@
 These are simply opportunities to share, learn, and exchange insights.
-Importantly, these sessions offer a chance to get to know one another and connect as fellow PhD students [~\cite{omeara2017, bernery2022}].
+Importantly, these sessions offer a chance to get to know one another and connect as fellow PhD students ~\cite{omeara2017, bernery2022}.
 
@@
 Schedule time to exchange written material and review that of your peers.
-An essential aspect of developing your writing is giving and receiving revisions [~\cite{guilford2001, reynolds2011}].
+An essential aspect of developing your writing is giving and receiving revisions ~\cite{guilford2001, reynolds2011}.
 Constructive feedback from peers may offer valuable insights on gaps in clarity, logic, and structure.
@@
 This practice encourages a collaborative approach to writing that builds confidence and resilience in both the reviewer and the reviewed.
-When giving feedback, use neutral language [~\cite{parsons2021}] and focus on writing components (structure, clarity, flow).
+When giving feedback, use neutral language ~\cite{parsons2021} and focus on writing components (structure, clarity, flow).
 If you are an expert in the content, detailed feedback can also be extremely valuable.
-Academics in general should be mindful of the historical barriers peer review imposes on minority groups and non-English speakers [~\cite{smith2023, saleh2024}], and work towards breaking this pattern.
-Furthermore, peer reviewing should be an integral part of academic life [~\cite{borja2024}].
+Academics in general should be mindful of the historical barriers peer review imposes on minority groups and non-English speakers ~\cite{smith2023, saleh2024}, and work towards breaking this pattern.
+Furthermore, peer reviewing should be an integral part of academic life ~\cite{borja2024}.
 Recognise peer review as a skill that can be practised, and a writing retreat could be the perfect place to begin if you are new to it.
@@
 While on a writing retreat, including recreational activities can enhance the overall experience.
-A writing retreat allows you to disconnect from the routine and responsibilities [~\cite{murray2013}], providing a unique opportunity to bond with your peers.
+A writing retreat allows you to disconnect from the routine and responsibilities ~\cite{murray2013}, providing a unique opportunity to bond with your peers.
 More rural places often have scenic landscapes, hospitable communities and limited distractions.
@@
 We organised creative writing prompts for each morning (see Rule 3) and gave casual presentations (Rule 7) about various topics, which were fun and engaging bonding activities.
-Furthermore, balancing downtime activities with writing blocks can lead to optimal productivity [~\cite{eardley2021, lyubykh2022}], enhancing the writing experience.
+Furthermore, balancing downtime activities with writing blocks can lead to optimal productivity ~\cite{eardley2021, lyubykh2022}, enhancing the writing experience.
 
@@
 Collectively, we experienced myriad benefits of writing retreats reported in the literature.
-These include feeling empowered as writers through supporting each other in a respectful and non-judgmental environment [~\cite{papen2018}], increasing our sense of belonging as PhD students [~\cite{omeara2017}] and decreasing our sense of isolation [~\cite{eardley2021}].
+These include feeling empowered as writers through supporting each other in a respectful and non-judgmental environment ~\cite{papen2018}, increasing our sense of belonging as PhD students ~\cite{omeara2017} and decreasing our sense of isolation ~\cite{eardley2021}.
 Most importantly, we achieved our writing goals at the retreats.
-Some of the references herein can help with the theoretical side of writing, but ultimately writing is practice and requires dedicated time [~\cite{peterson2018, grogan2021}].
+Some of the references herein can help with the theoretical side of writing, but ultimately writing is practice and requires dedicated time ~\cite{peterson2018, grogan2021}.
 By facilitating the planning and execution of writing retreats through these Ten Simple Rules, we hope to foster scholarly productivity and wellbeing in student cohorts.

### 2025-11-26 manuscript_ten-simple-rules-writing-retreat.tex
new version after internal edits and suggestions, plus uploading Supp Mat here

@@
 Consequently, many report a lack of time to think critically as a major challenge in academia ~\cite{menzies2007}, which too often leads researchers to deprioritise writing tasks ~\cite{boice1984, menzies2007}.
-Therefore, we must actively disengage from other tasks to fully engage with the act of writing ~\cite{murray2009, murray2013}.
+Therefore, to fully engage in the act of writing, many need to fully disengage from other tasks by carving out dedicated focus time ~\cite{murray2009, murray2013}.
 
-Writing retreats provide structured periods where academics dedicate time to focused writing ~\cite{mcgrail2006, murray2009}.
+Writing retreats provide structured periods where researchers dedicate time to focused writing ~\cite{mcgrail2006, murray2009}.
 These retreats offer practical opportunities to disconnect from daily work routines ~\cite{murray2013, tremblay2021}, which help researchers gain writing momentum and increase scholarly output ~\cite{mcgrail2006, kornhaber2016}.
 In addition, writing retreats foster a sense of community, promote wellbeing, and build self-confidence for academic writers ~\cite{kornhaber2016, eardley2021, tremblay2021}.
-Postgraduate students, in particular, highlight the value of these retreats in strengthening bonds among peers and appreciate the dedicated time and space to focus solely on writing ~\cite{kornhaber2016, papen2018, tremblay2021, bojovic2024}.
+Postgraduate students, in particular, highlight the value of these retreats in strengthening bonds among peers, obtaining and providing constructive feedback, and dedicating time and space to focus on writing ~\cite{kornhaber2016, papen2018, tremblay2021, bojovic2024}.
 
 As a cohort of PhD students in the Department of Marine Science at the University of Otago (Aotearoa/New Zealand), we organised a 5-day writing retreat at a remote field station in 2023.
-In New Zealand, PhDs follow research-only programs; as such, we do not participate in coursework that might promote group cohesion, and each student conducts research individually.
-In addition, given our diverse disciplines, field sites, and lab locations, we had felt a disconnection among us and a lack of community.
-Therefore, we initially planned the retreat to connect with our PhD cohort ~\cite{omeara2017, bernery2022}, but we also aimed to make it a productive work week ~\cite{mcgrail2006, kornhaber2016}.
-After returning from the retreat, we realised how much writing we had accomplished and the importance of the moments we had shared as postgrads going through similar experiences in various stages of our PhDs.
+In New Zealand, PhDs follow research-only programs; as such, we do not participate in coursework that might promote group cohesion, as each student leads their own research.
+In addition, given our diverse disciplines, field sites, and lab locations, there was a shared social disconnect within our programme.
+A writing retreat was planned to not only facilitate community building within our PhD cohort ~\cite{omeara2017, bernery2022}, but to also structure a productive week as we worked toward our dissertation goals ~\cite{mcgrail2006, kornhaber2016}.
+The retreat was a success both in terms of writing produced as well as connections built between peers sharing a similar PhD journey.
 The retreat's success motivated us to organise a second retreat the following year; the second retreat's success inspired a third.
@@
 Based on our experience, we present Ten Simple Rules for organising effective (student-led) writing retreats.
-Although the authors were all PhD students at the time of the retreats, these rules apply broadly to any research-oriented or academically-minded group.
+Although the authors were all PhD students at the time they participated in the retreats, these rules can be applied broadly to any research-oriented or academically-minded group.
 We outline steps to support the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and post- (Rule 10) writing retreat actions, but we do not cover writing techniques per se.
-Many resources on academic writing exist that we encourage you to explore (e.g.
-~\cite{turbek2016, hotaling2020, carter2020, cargill2021}), including articles in this series ~\cite{weinberger2015, mensh2017, peterson2018}---or you may simply choose to write a boring contribution instead ~\cite{sand2007}.
+Many helpful resources on academic writing exist (e.g.
+~\cite{turbek2016, hotaling2020, carter2020, cargill2021}), including articles in this series ~\cite{weinberger2015, mensh2017, peterson2018}.
 
@@
 A change of scenery inspires productivity ~\cite{murray2013}, whether you organise the writing retreat locally or as a multi-day trip.
+When assessing suitability of the location, also consider internet access, heating/cooling, sleeping arrangements, needed accommodations for participants with particular needs, and workspace infrastructure (power points, seats, tables, etc.).
+Taking advantage of institutional resources simplifies the planning process and enhances the success of your writing retreat.
 Local facilities, such as meeting rooms and other bookable spaces on campus, offer accessible options.
@@
 For example, many biology and ecology programmes maintain field stations (for a global list, see ~\cite{tydecks2016}).
-By using university-owned facilities, students keep retreat costs manageable.
+By using university-owned facilities, students can likely keep retreat costs manageable.
+
 Field stations often remain underutilized during certain periods, so retreats during these times can be used to promote year-round upkeep of facilities.
@@
 
-When evaluating a potential site for a retreat, it is important to review available amenities both on-site and nearby.
-Also, make sure the space accommodates people with physical disabilities or impairments.
-We made use of our department's marine field station in Oban, Rakiura/Stewart Island (a small island off the southern coast of New Zealand's South Island).
+For our writing retreat, we made use of our department's marine field station in Oban, Rakiura/Stewart Island (a small island off the southern coast of New Zealand's South Island).
 This location suited us well---everything we needed was within walking distance.
-Because this is a small community (around 400 year-round residents) abutting a national forest, we had limited distractions, straightforward options for restaurants and groceries, and ample access to walking tracks and beach strolls.
-When assessing suitability of the location, also consider internet access, heating/cooling, sleeping arrangements, and workspace infrastructure (power points, seats, tables, etc.).
-Taking advantage of institutional resources simplifies the planning process and enhances the success of your writing retreat.
+Because this is a small community (around 400 year-round residents) abutting a national forest, we had limited distractions, straightforward options for restaurants and groceries, and ample access to walking tracks and beaches.
 
@@
 Clearly define the capacity and target audience to plan details of the retreat effectively.
-Your options for accommodation (see Rule 1) may determine your maximum number of participants.
+Your options for facilities (see Rule 1) may determine your maximum number of participants.
 In our case, we hosted all retreats with 10 participants; we found this number worked well for the size of our facility and the length of the retreat.
@@
 
-Create a schedule to structure your writing retreat, as this will be key to its success ~\cite{murray2009, tremblay2021}.
+Create a schedule for your writing retreat, as it provides the structure for focused writing and thinking time ~\cite{murray2009, tremblay2021}.
 Use available resources to guide your planning (e.g.
@@
 At our university, we collaborated with the Higher Education Development Centre, which provided guidance on designing productive days and structuring the week for our first and second retreats.
-We encourage you to take advantage of such services when available.
 
-We travelled to the accommodation on Sunday and returned the following Saturday, allowing Monday through Friday to remain fully structured for retreat activities.
+The first and last day of the schedule were dedicated to travelling to and from the retreat location, allowing for the five days in between to remain fully structured for the retreat activities.
 We built in extra time on travel days to settle in, pack up, shop for groceries, and explore the area.
-Upon arrival, we held a welcome discussion to establish ground rules and ensure everyone felt comfortable (see Rule 5).
-We themed each workday to provide guidance and structure for participants (Table~\ref{table1}), while allowing each individual to pursue their own goals.
+Upon arrival, we held a welcome discussion to establish ground rules to ensure everyone felt comfortable and had a shared understanding of expectations (see Rule 5).
+Each work day was themed to provide guidance and structure for participants (Table~\ref{table1}), while allowing each individual to pursue their own goals.
 We also incorporated optional social activities throughout the week (see Rule 9), such as a pub quiz, wildlife tours, game night, and local events.
@@
 \centering
-\caption{{\bf Examples of daily themes over a 5-day writing retreat. Note that days 1 and 7 were organised for travel to and settling in, and packing up and travelling back, respectively.}}
+\caption{{\bf Examples of daily themes over a 5-day writing retreat. Note that days 1 and 7 were organised for travel.}}
 \begin{tabular}{p{1in}p{2in}p{2in}}
@@
 
-Our daily schedule typically ran from 9:00 AM to 6:00 PM and was shared in advance so participants knew when to expect focused work sessions ~\cite{tremblay2021} (see also \nameref{S1_Appendix}).
-We designated a different facilitator each day to lead discussions and manage writing blocks (60--120 minutes) and breaks (30--60 minutes).
+Our daily schedule typically ran from 9:00 AM to 6:00 PM and was shared in advance so participants knew what to expect during the week ~\cite{tremblay2021} (see also \nameref{S1_Appendix}).
+A different facilitator was designated each day to lead discussions and manage writing blocks (60--120 minutes) and breaks (30--60 minutes).
 Each morning began with a 15-minute writing session using prompts that ranged from reflective writing to humorous research titles (see Rule 9).
-These sessions helped rev up our writing engines for the day and offered a fun, low-pressure opportunity to share writing.
-Each evening, we closed the day with a reflective discussion on how the day went (with an `accountability partner'; see Rule 5 in ~\cite{peterson2018}), a shared dinner, and presentations (see Rule 7).
+These sessions helped prime our writing engines for the day and offered a fun, low-pressure opportunity to share writing.
+Each evening, the day was closed with a reflective discussion on how the day went (with an `accountability partner'; see Rule 5 in ~\cite{peterson2018}), a shared dinner, and presentations (see Rule 7).
 
@@
 This meeting gives participants a chance to meet, finalise logistics, set goals, and raise questions or concerns.
-Use this time to build or reinforce a respectful and accountable group dynamic (see Rules 5 and 10).
-Collaboratively fine-tune logistics and schedules to clarify plans and foster shared ownership.
-Most importantly, a pre-retreat meeting allows participants to identify or workshop specific writing goals with time to prepare relevant resources (literature, data analysis, input from academic supervisors) so that non-writing activities do not become distractions during the initial days of the retreat.
+This time can be used to build or reinforce a respectful and accountable group dynamic (see Rules 5 and 10), and collaboratively fine-tune logistics and schedules.
+Most importantly, a pre-retreat meeting allows participants to identify or set specific writing goals while there is still time to prepare relevant resources (literature, data analysis, input from academic supervisors).
+A pre-retreat meeting can help to prompt participants to organise themselves before leaving so that non-writing activities, like gathering references, do not become distractions during the initial days of the retreat.
 
@@
 For reference, during our 5-day retreats, participants averaged 1.2 pages or 550 words per day, working on everything from detailed revisions to drafting full manuscripts.
-Identifying a specific writing target ahead of time reduces decision-making friction at the retreat and maximises productive preparation.
+By identifying a specific writing target ahead of time, friction related to making decisions is reduced during the retreat with goals clear in mind.
 Sharing goals also helps participants learn about each other's research and assess whether goals are realistic.
 
-Hold the pre-retreat meeting within two weeks of departure.
+The pre-retreat meeting should be held within two weeks of departure to support timely preparation while preserving participant momentum.
 This timeframe supports goal-oriented preparation while preserving participant momentum.
-If possible, host the meeting in person in a casual setting.
-If attendees join from different campuses, a video call works as well.
-Additional objectives may include selecting writing prompts, presentation topics, icebreakers, roles (e.g.
-meal coordination, daily facilitator), or ideas for social activities.
-
-If you cannot host a pre-retreat meeting, incorporate these objectives (introductions, goal-setting, and planning) into the first day of the retreat, ideally following the discussion of ground rules (see Rule 5).
-However, doing so may reduce writing time and diminish early momentum.
+If possible, a casual, in-person meeting is preferred to provide a comfortable setting for participants to meet and connect.
+Additional objectives may include selecting writing prompts, presentation topics, icebreakers, roles (e.g., meal coordination, daily facilitator), or ideas for social activities.
 
@@
 Establish clear ground rules and expectations at the start of the retreat.
-These include shared expectations for the group, such as setting quiet hours, assigning household responsibilities (e.g.
-cooking and cleaning), and setting guidelines for breaks.
-You can find the document we used as the basis for our ground rules in \nameref{S2_Appendix}.
+These include shared expectations for the group, such as setting quiet hours, assigning household responsibilities (e.g., cooking and cleaning), and setting guidelines for breaks.
 Ground rules should also set the tone of the retreat, creating a balanced foundation for respect, productivity, and enthusiasm.
@@
 For example, not all participants were native English speakers and may find full days of writing and discussion in English more tiring.
-As a result, we designated writing blocks as optional---if participants need a longer break or feel they would benefit from refreshing their mind with a forest walk or gym session, they should do so without judgement.
-Continuing with our guideline of respect, we also made it clear in our opening meeting that while group discussions and feedback are useful, we only tolerate constructive criticism.
+As a result, we designated writing blocks as optional---if participants needed a longer break or felt they would benefit from a forest walk or gym session, they were free to do so without judgement.
+Continuing with our guideline of respect, we also made it clear in our opening meeting that while group discussions and feedback are useful, we would only tolerate constructive criticism.
 Together, clear expectations, a set schedule, and an overall attitude of respect set the tone for a successful week of writing.
+
+The document we used as the basis for our ground rules is in \nameref{S2_Appendix}.
 
@@
 This is where individual goals (Rule 4) come into play.
-Knowing what you are working towards, and having a plan to get there, provides the bandwidth needed to put words on the page and have productive writing blocks.
+Knowing what you are working towards, and having a plan to get there, provides the structure needed to put words on the page and have productive writing blocks.
 
@@
 Some authors, such as Ernest Hemingway, have even suggested to leave a sentence, paragraph, or idea unfinished at the end of a writing day so they have something easy to start with during the next writing block ~\cite{hemingway1935}.
-Assess what works best for your writing practice and reevaluate regularly to ensure that strategy is still serving you ~\cite{peterson2018, grogan2021}.
+Assess what works best for your writing practice and reevaluate regularly to ensure that a strategy is still serving you ~\cite{peterson2018, grogan2021}.
 
@@
 A writing retreat gathers your academic peers in a collaborative environment with minimal external distractions.
-It presents a unique opportunity to garner advice and perspectives on written work (see also Rule 8) or any work-in-progress that participants might be developing.
-
-Undertaking any large-writing project (e.g.
-a PhD thesis) is a daunting and varying experience for everyone.
-It is easy to become so engrossed in your work that you forget you are surrounded by people in a similar situation ~\cite{wilson2019}.
-A writing retreat provides the opportunity to share your own research or discuss scientific/academic topics amongst your peers.
-While we found this occurred naturally throughout the retreats, especially within groups of similarly focused researchers, we also allocated a small block of time each day (see Rule 3) specifically for students to share a talk or lead a discussion with the other attendees.
-The casual, non-judgemental environment allowed presenting students a chance to practice a seminar talk, gain insight on methodologies, analyses and results, or simply share `tips and tools of the trade.'
-
+It presents a unique opportunity to garner advice and perspectives on written work (see also Rule 8) or any work-in-progress that participants may be developing.
 We suggest scheduling these discussions after writing blocks are complete for the day and spreading them out over the retreat.
@@
 These are simply opportunities to share, learn, and exchange insights.
+
+While we found academic discussions materalised naturally throughout the retreats, especially within groups of similarly focused researchers, we also allocated a small block of time each day (see Rule 3) specifically for students to share a talk or lead a discussion with the other attendees.
+The casual, non-judgemental environment allowed presenting students a chance to practice a seminar talk, gain insight on methodologies, analyses and results, or simply share `tips and tools of the trade.' Undertaking any large-writing project (e.g., a PhD thesis) is a daunting and varying experience for everyone.
+It is easy to become so engrossed in your work that you forget you are surrounded by people in a similar situation ~\cite{wilson2019}.
+A writing retreat provides the opportunity to share your own research or discuss scientific/academic topics amongst your peers.
 Importantly, these sessions offer a chance to get to know one another and connect as fellow PhD students ~\cite{omeara2017, bernery2022}.
@@
 
-Schedule time to exchange written material and review that of your peers.
-An essential aspect of developing your writing is giving and receiving revisions ~\cite{guilford2001, reynolds2011}.
+Schedule time to exchange written material and review the work of other participants.
+An essential aspect of developing your writing is critical review both in terms of giving and receiving feedback as well as through editing your own work ~\cite{guilford2001, reynolds2011}.
 Constructive feedback from peers may offer valuable insights on gaps in clarity, logic, and structure.
@@
 This practice encourages a collaborative approach to writing that builds confidence and resilience in both the reviewer and the reviewed.
-When giving feedback, use neutral language ~\cite{parsons2021} and focus on writing components (structure, clarity, flow).
+When giving feedback, use neutral language ~\cite{parsons2021} and focus on writing components for example, structure, clarity, and flow.
 If you are an expert in the content, detailed feedback can also be extremely valuable.
-Academics in general should be mindful of the historical barriers peer review imposes on minority groups and non-English speakers ~\cite{smith2023, saleh2024}, and work towards breaking this pattern.
-Furthermore, peer reviewing should be an integral part of academic life ~\cite{borja2024}.
-Recognise peer review as a skill that can be practised, and a writing retreat could be the perfect place to begin if you are new to it.
+
+Peer review is an integral part of academic life ~\cite{borja2024}, that needs to be practised.
+Furthermore, researchers should be mindful of the historical barriers peer review imposes on minority groups and non-English speakers ~\cite{smith2023, saleh2024}, and work towards breaking this pattern.
+A writing retreat is a friendly, safe space to talk about these barriers (Rule 7) and become more comfortable with the peer-review process in general while practising writing critical, neutral, and encouraging reviews.
 
@@
 Have fun and build connections with your peers.
-While on a writing retreat, including recreational activities can enhance the overall experience.
-A writing retreat allows you to disconnect from the routine and responsibilities ~\cite{murray2013}, providing a unique opportunity to bond with your peers.
-More rural places often have scenic landscapes, hospitable communities and limited distractions.
-While this may restrict choices for activities, it can lead to higher quality experiences where you can get to know local businesses and each other on a more personal level.
+While on a writing retreat, recreational activities can enhance the overall experience.
+Furthermore, balancing downtime activities with writing blocks can lead to optimal productivity ~\cite{eardley2021, lyubykh2022}, improving the writing experience.
+A writing retreat allows you to disconnect from regular routines and responsibilities ~\cite{murray2013}, providing a unique opportunity to bond with your peers.
+Rural places often have scenic landscapes, hospitable communities and limited distractions.
+While this may restrict choices for activities, it can lead to higher-quality experiences where you can get to know local businesses and each other on a more personal level.
 Before embarking on a writing retreat, research what is unique about the location and take advantage of what may be on offer.
-We organised creative writing prompts for each morning (see Rule 3) and gave casual presentations (Rule 7) about various topics, which were fun and engaging bonding activities.
-Furthermore, balancing downtime activities with writing blocks can lead to optimal productivity ~\cite{eardley2021, lyubykh2022}, enhancing the writing experience.
 
-\section*{Rule 10: Posterity—gather feedback and demonstrate value}
+While based at our university's field station, we had opportunities to explore the nearby Ulva Island (a protected area), participate in the locally famous bar trivia, and go on aurora hunts and kiwi spotting at night.
+In addition, we organised creative writing prompts each morning (see Rule 3; some of them hosted at the local cafe), which created fun moments to start the day.
+The casual presentations (Rule 7) about various topics were fun and engaging bonding activities as well.
+
+\section*{Rule 10: Posterity---gather feedback and demonstrate value}
 
@@
 In addition, publicly acknowledging the retreat in your writing and presentations demonstrates its academic value.
-Include it in the acknowledgements of manuscripts and theses worked on during the retreat.
+Participants are encouraged to include the retreat in the acknowledgements of manuscripts and theses worked on during the retreat.
 Sharing this impact helps advocate for continued support, especially in institutions where publication output informs resource allocation.
 Maintaining a list of retreat-supported outputs can strengthen future proposals (see Rule 2) and make a compelling case for funding.
-Indeed, demonstrating the immense value of this experience may spur on support from leadership to fund retreats for students in other departments or institutions.
+Indeed, demonstrating the immense value of this experience may spur future support from leadership to fund retreats for students in other departments or institutions.
 
@@
 
-Collectively, we experienced myriad benefits of writing retreats reported in the literature.
-These include feeling empowered as writers through supporting each other in a respectful and non-judgmental environment ~\cite{papen2018}, increasing our sense of belonging as PhD students ~\cite{omeara2017} and decreasing our sense of isolation ~\cite{eardley2021}.
+Collectively, we experienced many of the benefits of writing retreats reported in the literature.
+These included feeling empowered as writers through supporting each other in a respectful and non-judgmental environment ~\cite{papen2018}, increasing our sense of belonging as PhD students ~\cite{omeara2017} and decreasing our sense of isolation ~\cite{eardley2021}.
 Most importantly, we achieved our writing goals at the retreats.
 Some of the references herein can help with the theoretical side of writing, but ultimately writing is practice and requires dedicated time ~\cite{peterson2018, grogan2021}.
-By facilitating the planning and execution of writing retreats through these Ten Simple Rules, we hope to foster scholarly productivity and wellbeing in student cohorts.
+By facilitating the planning and execution of writing retreats through these Ten Simple Rules, we hope to foster scholarly productivity and wellbeing in student and early career cohorts.
 Happy writing!
@@
 \section*{Acknowledgments}
-This manuscript is a contribution conceptualised, led, and written by PhD students, who attended writing retreats supported by the Department of Marine Science of the University of Otago, Aotearoa/New Zealand.
+This manuscript is a contribution conceptualised, led, and written by PhD students, who attended and organised writing retreats supported by the Department of Marine Science of the University of Otago, Aotearoa/New Zealand.
 We hope writing retreats stay as a trend in our Department, fostering collaboration and friendship among students and promoting quality science.

### 2025-12-07 manuscript_ten-simple-rules-writing-retreat.tex
fixed typos and minor adjustments before ms submission

@@
 % Current address notes
-\textcurrency Current Address: Marine Mammal Institute, Department of Fisheries, Wildlife and Conservation Sciences, Oregon State University, Newport, OR, USA
+\textcurrency Current Address: Geospatial Ecology of Marine Megafauna Laboratory, Marine Mammal Institute, Oregon State University, Newport, OR, USA \& Department of Fisheries, Wildlife, and Conservation Sciences, Oregon State University, Corvallis, OR, USA
 
@@
 In New Zealand, PhDs follow research-only programs; as such, we do not participate in coursework that might promote group cohesion, as each student leads their own research.
-In addition, given our diverse disciplines, field sites, and lab locations, there was a shared social disconnect within our programme.
+In addition, our diverse disciplines, field sites, and lab locations present challenges in maintaining social cohesion within our programme.
 A writing retreat was planned to not only facilitate community building within our PhD cohort ~\cite{omeara2017, bernery2022}, but to also structure a productive week as we worked toward our dissertation goals ~\cite{mcgrail2006, kornhaber2016}.
@@
 We outline steps to support the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and post- (Rule 10) writing retreat actions, but we do not cover writing techniques per se.
-Many helpful resources on academic writing exist (e.g.
-~\cite{turbek2016, hotaling2020, carter2020, cargill2021}), including articles in this series ~\cite{weinberger2015, mensh2017, peterson2018}.
+Many helpful resources on academic writing exist (e.g., ~\cite{turbek2016, hotaling2020, carter2020, cargill2021}), including articles in this series ~\cite{weinberger2015, mensh2017, peterson2018}.
 
@@
 There may be external funding available to provide supplemental support if internal money is not available to cover the entire costs of the retreat.
-This funding could be sought from industry and community organisations (e.g.
-Lions and Rotary Clubs), or through fundraisers.
+This funding could be sought from industry and community organisations (e.g., Lions and Rotary Clubs), or through fundraisers.
 A well-developed proposal lays a solid foundation for organising the retreat and securing the support needed for success.
@@
 Create a schedule for your writing retreat, as it provides the structure for focused writing and thinking time ~\cite{murray2009, tremblay2021}.
-Use available resources to guide your planning (e.g.
-~\cite{tremblay2021}), including this guide (see also Appendices).
+Use available resources to guide your planning (e.g., ~\cite{tremblay2021}), including this guide (see also Appendices).
 At our university, we collaborated with the Higher Education Development Centre, which provided guidance on designing productive days and structuring the week for our first and second retreats.
@@
 We suggest that participants arrive at the pre-retreat meeting with a preliminary goal to share with the group.
-Goals may be quantitative (e.g.
-word count, page number) or qualitative (e.g.
-specific sections or revisions).
+Goals may be quantitative (e.g., word count, page number) or qualitative (e.g., specific sections or revisions).
 For reference, during our 5-day retreats, participants averaged 1.2 pages or 550 words per day, working on everything from detailed revisions to drafting full manuscripts.
@@
 The pre-retreat meeting should be held within two weeks of departure to support timely preparation while preserving participant momentum.
-This timeframe supports goal-oriented preparation while preserving participant momentum.
 If possible, a casual, in-person meeting is preferred to provide a comfortable setting for participants to meet and connect.
@@
 Discussions were informal, and we aimed not to exceed 20 minutes per session.
-It is important to emphasise the casualness of such an exercise---students should not be using their time for writing anxiously preparing a conference-level presentation.
+It is important to emphasise the casualness of such an exercise---students should not anxiously prepare a conference-level presentation at the expense of their writing time.
 These are simply opportunities to share, learn, and exchange insights.
 
-While we found academic discussions materalised naturally throughout the retreats, especially within groups of similarly focused researchers, we also allocated a small block of time each day (see Rule 3) specifically for students to share a talk or lead a discussion with the other attendees.
+While we found academic discussions materialised naturally throughout the retreats, especially within groups of similarly focused researchers, we also allocated a small block of time each day (see Rule 3) specifically for students to share a talk or lead a discussion with the other attendees.
 The casual, non-judgemental environment allowed presenting students a chance to practice a seminar talk, gain insight on methodologies, analyses and results, or simply share `tips and tools of the trade.' Undertaking any large-writing project (e.g., a PhD thesis) is a daunting and varying experience for everyone.
@@
 This practice encourages a collaborative approach to writing that builds confidence and resilience in both the reviewer and the reviewed.
-When giving feedback, use neutral language ~\cite{parsons2021} and focus on writing components for example, structure, clarity, and flow.
+When giving feedback, use neutral language ~\cite{parsons2021} and focus on writing components, such as structure, clarity, and flow.
 If you are an expert in the content, detailed feedback can also be extremely valuable.
